%*******************************************************
% Declaration
%*******************************************************

\ifdefined\ThesisLanguageIsEnglish
    \chapter*{Declaration}
    \thispagestyle{empty}

    Hereby I assure that I have written the presented practical training report independently and without any third party help and that I have not used any other tools or resources than the ones mentioned/referred to in the report.
    \medskip

    \noindent
    Any parts of the report that have been taken from other published or yet unpublished works in terms of their wording or meaning are thoroughly marked with an indication of the source.
    \medskip

    \noindent
    All figures in this report have been drawn by myself or are clearly attributed with a reference.
    \medskip

    \noindent
    This work has not been published or presented to any other examination authority before.
    I am aware of the importance of the affidavit and the consequences under examination
    law as well as the criminal consequences of an incorrect or incomplete affidavit.
    \medskip


\else
    \chapter*{Eidesstattliche Erklärung}
    \thispagestyle{empty}
    Ich versichere hiermit, dass ich die vorliegende Arbeit selbständig verfasst und keine anderen als die im Literaturverzeichnis angegebenen Quellen benutzt habe.
    \medskip

    \noindent
    Alle Stellen, die wörtlich oder sinngemäß aus veröffentlichten oder noch nicht veröffentlichten Quellen entnommen sind, sind als solche kenntlich gemacht.
    \medskip

    \noindent
    Die Zeichnungen oder Abbildungen in dieser Arbeit sind von mir selbst erstellt worden oder mit einem entsprechenden Quellennachweis versehen.
    \medskip

    \noindent
    Diese Arbeit ist in gleicher oder ähnlicher Form noch bei keiner anderen Prüfungsbehörde eingereicht worden.
    \bigskip
\fi

Ho Chi Minh City, \ThesisDeliveryDate

\smallskip


\hfill \makebox[10cm]{ } \\
\vspace*{-1ex}
\begin{flushright}
    \begin{tabular}{m{5cm}}
        \hline
        \vspace{0.3cm}
        \centering \textbf{\myName} \\
    \end{tabular}
\end{flushright}
%\textbf{\myName}