\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
\label{ch:Abstract}

Traffic signal control at the network level constitutes a canonical multi-agent decision problem in which inter-agent coordination can substantially reduce urban congestion. Graph-based multi-agent reinforcement learning (MARL) methods, which treat signalized intersections as nodes in a communication graph, have shown promise in synthetic environments but remain underexplored on real-world German road networks characterized by irregular topologies and heterogeneous phase structures.

This thesis proposes FGS (\textbf{F}ramework for \textbf{G}raph-aware \textbf{S}ignal control), a modular three-module pipeline for MARL-based traffic signal control that decouples the local encoder, graph communication layer, and RL algorithm as independently configurable design choices. Three successive FGS instantiations — FGSv1, FGSv2, and FGSv3 — explore this design space through an ablation-driven diagnosis-and-repair cycle: each generation is motivated by a specific module-interface failure identified in evaluation of the preceding one. FGSv2 preserves per-phase information across the encoder–communication boundary via action tokens. FGSv3 replaces competition-derived communication signals with demand-aware embeddings and introduces a factored neighborhood critic ($O(d_c)$ independent of~$N$), enabling both SAC and PPO to complete training at the largest evaluated scenario. The framework is implemented on SUMO, SUMO-RL, and RLlib under the CTDE paradigm.

All six German RESCO benchmark scenarios spanning three scale tiers — Single-intersection (cologne1, ingolstadt1), Corridor (cologne3, ingolstadt7), and Regional (cologne8, ingolstadt21) — are evaluated against seven baselines: fixed-time control, max-pressure, IDQN, IPPO, independent SAC, FRAP, and CoLight. Performance is reported on mean waiting time, mean delay, mean trip duration, and throughput.

A component-level ablation across all eight design-space combinations establishes three design guidelines. First, MLP encoders outperform FRAP-based encoders on German intersection heterogeneity by 5.2--62$\times$ depending on scenario; FRAP's phase-competition priors are structurally mismatched with German variable-phase layouts. Second, GATv2 dynamic attention reduces mean waiting time by 66\% over GAT at Regional scale ($N=21$; 88.6\,s vs.\ 261.2\,s) while providing negligible benefit at smaller scale. Third, SAC outperforms PPO at small-to-moderate scale (21\% reduction on cologne8) but requires the FGSv3 factored critic to complete training at $N=21$.

FGSv3-PPO achieves 22.9\,s mean waiting time on ingolstadt21, outperforming CoLight (120.8\,s, $-81\%$) and all other tested baselines. FGS(MLP+GATv2+SAC) attains 4.93\,s on the Regional cologne8 scenario, outperforming CoLight (11.0\,s) by 55\%. These results demonstrate that the FGS pipeline is an effective and analytically informative framework for characterizing MARL-TSC design on German road networks.

\medskip
\noindent\textbf{Keywords:} multi-agent reinforcement learning, traffic signal control, graph attention network, soft actor-critic, FRAP encoder, RESCO benchmark, decentralized execution, German traffic scenarios

\chapter*{Kurzfassung}
\addcontentsline{toc}{chapter}{Kurzfassung}
\label{ch:Kurzfassung}

Die Verkehrssignalsteuerung auf Netzwerkebene stellt ein kanonisches Mehrfach-Agenten-Entscheidungsproblem dar, bei dem eine koordinierte Abstimmung zwischen Agenten den städtischen Verkehrsstau erheblich verringern kann. Graphbasierte Methoden des Multi-Agenten Reinforcement Learning (MARL), die Lichtsignalanlagen als Knoten in einem Kommunikationsgraphen modellieren, haben in synthetischen Umgebungen vielversprechende Ergebnisse gezeigt, wurden jedoch für reale deutsche Straßennetze mit unregelmäßigen Topologien und heterogenen Phasenstrukturen kaum untersucht.

Diese Arbeit schlägt FGS (\textbf{F}ramework for \textbf{G}raph-aware \textbf{S}ignal control) vor, eine modulare Drei-Modul-Pipeline für MARL-basierte Verkehrssignalsteuerung, die lokalen Kodierer, Graphkommunikationsschicht und RL-Algorithmus als unabhängig konfigurierbare Entwurfsentscheidungen entkoppelt. Drei aufeinanderfolgende FGS-Instanziierungen — FGSv1, FGSv2 und FGSv3 — erkunden diesen Entwurfsraum durch einen ablationsgetriebenen Diagnose-und-Reparatur-Zyklus: Jede Generation adressiert einen spezifischen Modul-Schnittstellenfehler, der in der Evaluation der vorherigen Generation identifiziert wurde. FGSv2 bewahrt phasenspezifische Informationen an der Kodierer-Kommunikations-Schnittstelle mittels Aktions-Token. FGSv3 ersetzt wettbewerbsbasierte Kommunikationssignale durch nachfrageorientierte Einbettungen und führt einen faktorierten Nachbarschaftskritiker ein ($O(d_c)$ unabhängig von~$N$), der SAC- und PPO-Training im größten evaluierten Szenario ermöglicht. Das Framework ist auf Basis von SUMO, SUMO-RL und RLlib nach dem CTDE-Paradigma implementiert.

Alle sechs deutschen RESCO-Benchmark-Szenarien in drei Skalierungsstufen — Einzelkreuzung (cologne1, ingolstadt1), Korridor (cologne3, ingolstadt7) und Regional (cologne8, ingolstadt21) — werden gegen sieben Basisverfahren evaluiert: Festzeitsteuerung, Maximaldruck, IDQN, IPPO, unabhängiges SAC, FRAP und CoLight.

Eine komponentenweise Ablation über alle acht Entwurfsraumkombinationen liefert drei Entwurfsrichtlinien. Erstens übertreffen MLP-Kodierer FRAP-basierte Kodierer auf deutschen Kreuzungen um den Faktor 5,2--62 je nach Szenario, da FRAPs Phasenwettbewerbs-Annahmen strukturell mit deutschen variablen Phasenstrukturen inkompatibel sind. Zweitens reduziert GATv2-Aufmerksamkeit die mittlere Wartezeit im Regionalmaßstab ($N=21$) um 66\,\% gegenüber GAT (88,6\,s vs.\ 261,2\,s), bietet jedoch bei kleinerem Maßstab keinen messbaren Vorteil. Drittens übertrifft SAC PPO bei kleinen bis mittleren Netzwerken (21\,\% Reduktion auf cologne8), erfordert jedoch den faktorierten Kritiker von FGSv3, um das Training bei $N=21$ abzuschließen.

FGSv3-PPO erreicht 22,9\,s mittlere Wartezeit auf ingolstadt21 und übertrifft damit CoLight (120,8\,s, $-81\,\%$) sowie alle übrigen Basisverfahren. FGS(MLP+GATv2+SAC) erzielt 4,93\,s auf cologne8 und übertrifft CoLight (11,0\,s) um 55\,\%. Diese Ergebnisse belegen, dass die FGS-Pipeline ein effektives und analytisch aufschlussreiches Framework für die Charakterisierung von MARL-TSC-Entwurfsentscheidungen auf deutschen Straßennetzen darstellt.

\medskip
\noindent\textbf{Schlüsselwörter:} Multi-Agenten Reinforcement Learning, Verkehrssignalsteuerung, Graphaufmerksamkeitsnetzwerk, Soft Actor-Critic, FRAP-Kodierer, RESCO-Benchmark, dezentralisierte Ausführung, deutsche Verkehrsszenarien
