\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
\label{ch:Acknowledgements}

This thesis would not have been possible without the guidance, support, and encouragement of many people throughout my journey.

First and foremost, I would like to express my deepest gratitude to my first supervisor, Dr. Nguyen Tuan Cuong. Over the past years, through both my thesis work and my experience as his teaching assistant, I have learned a great deal from his guidance, patience, and dedication. His regular meetings, thoughtful advice, and timely support helped me keep this thesis on the right path and overcome many challenges along the way.

I am also sincerely grateful to Prof. Dr. Javad Ghofrani for his valuable supervision and insightful feedback throughout this research. His perspectives helped me reflect more deeply on the direction of the work and improve the quality of this thesis.

I would like to thank Mr. Ton That Nhat Minh, my friend and fellow researcher, for accompanying me through this research journey. I truly appreciate the long discussions, the late nights, and the shared effort we put into developing ideas, baselines, methods, and analyses. Working with him made this challenging process more meaningful, and I am grateful for both his dedication and his companionship.

My sincere thanks also go to Asst. Prof. Pham Duc Thinh and Dr. Le Trung Hoang at the Center for AI Research, VinUniversity. Although their guidance was not directly tied to this thesis, their mentorship in another traffic control project has greatly shaped the way I approach research. Their insights have helped me take important steps toward becoming a more independent researcher.

I would also like to thank my friends, relatives, and family for their constant emotional support throughout this journey. In particular, I am deeply grateful to Mr. Dang Tran Long and Ms. Dang Phuong Mai for being with me through the ups and downs, especially during the moments when I doubted myself. Their encouraging words, care, and presence gave me strength when I needed it most.

Finally, I would like to quietly thank myself for not giving up. This journey came with pain, pressure, and many challenges, but I am grateful that I kept going, one step at a time.
