\chapter{Introduction}
\label{ch:Introduction}

\section{Background and Motivation}
\label{sec:Introduction_BackgroundAndMotivation}

Urban traffic congestion is among the most persistent inefficiencies in modern infrastructure. In cities worldwide, the time lost to congestion increases vehicle emissions, raises logistics costs, and degrades quality of life for millions of daily commuters. Germany, with one of the highest vehicle densities in Europe, experiences significant peak-hour delays in its major urban centers; Cologne and Ingolstadt — the two cities whose road networks are evaluated in this thesis — are representative of the medium-density European urban environment where signal timing directly determines the throughput and fairness of the road network.

Traffic signal control (TSC) is the primary lever through which urban traffic is managed at the intersection level. A well-timed signal plan reduces average waiting time, balances queue lengths across directions, and enables coordinated flow along arterials. Conversely, poorly timed signals generate spillback — queues that extend upstream and block intersections they do not serve — which can cascade into network-wide congestion. The challenge is that optimal signal timing depends on real-time demand conditions that are neither stationary nor predictable from historical data alone. Fixed-time plans, designed for average demand, leave substantial throughput gains on the table during off-peak periods and fail under demand spikes caused by incidents or events.

The response to these limitations has followed two broad tracks. The first track, represented by adaptive control systems such as SCOOT \citep{hunt1981scoot} and SCATS \citep{lowrie1990scats}, augments traditional signal engineering with real-time feedback from loop detectors. These systems have demonstrated meaningful improvements in arterial delay but require dense sensing infrastructure, expert parameter calibration, and remain constrained by pre-defined phase sequences. The second track, initiated by early reinforcement learning (RL) work \citep{wiering2000multiagent} and greatly expanded by deep RL \citep{wei2018intellilight,wei2019presslight}, treats phase selection as a sequential decision problem that an agent can learn to solve from simulation experience. This approach requires no explicit traffic model and, in principle, adapts to any demand pattern within the distribution encountered during training.

The extension from single-intersection RL to multi-agent RL (MARL) for network-level control introduces a further challenge: agents must not only learn individually effective phase policies, but also produce collectively coherent behavior that reduces network-wide delay. Independent agents, each optimizing its own local reward, ignore the upstream and downstream consequences of their decisions \citep{chen2020toward,ault2021reinforcement}. Graph-based communication — where each intersection aggregates information from its neighbors over the road graph — provides a mechanism for coupling agent decisions without requiring centralized coordination at execution time \citep{wei2019colight}.


\section{Problem Statement}
\label{sec:Introduction_ProblemStatement}

Despite significant progress in RL-based TSC, three practical gaps remain unaddressed in the published literature.

First, the dominant evaluation environments in MARL-TSC research are synthetic grid networks or Chinese urban scenarios (Hangzhou, Jinan). The German RESCO benchmark \citep{ault2021reinforcement} provides real-world demand models for Cologne and Ingolstadt and has been adopted as a standard evaluation protocol, but few MARL-TSC methods — particularly those using graph-based communication — have been evaluated on it. Whether coordination mechanisms that work on regular Chinese grids transfer to the irregular topologies and heterogeneous phase structures of German networks is an open empirical question.

Second, the RL algorithms used in MARL-TSC are almost exclusively value-based (DQN and its variants) or on-policy (PPO). Soft Actor-Critic \citep{haarnoja2018soft}, which offers off-policy sample efficiency and entropy-regularized exploration, has not been systematically evaluated in a graph-aware multi-agent setting for traffic control. SAC's entropy bonus may be particularly beneficial in multi-agent environments, where diverse policies reduce the risk that agents converge to coordination patterns that perform well in training scenarios but fail under distribution shift.

Third, the design space of MARL-TSC architectures — encoder type, communication mechanism, and RL algorithm — has not been systematically mapped or ablated on German road networks. Whether phase-aware encoding complements graph attention, whether entropy regularization helps or hinders convergence in multi-agent German scenarios, and at what network scales each component provides consistent benefit are open empirical questions. No published method treats these three dimensions as independently configurable choices and evaluates their interactions on the German RESCO benchmark.

This thesis addresses all three gaps through the FGS (\textbf{F}ramework for \textbf{G}raph-aware \textbf{S}ignal control): a modular three-module pipeline that decouples encoder, graph communication, and RL algorithm as independent design choices, implemented across three successive generations (FGS, FGSv2, FGSv3) and evaluated on six German RESCO scenarios spanning three scale tiers: Single-intersection (cologne1, ingolstadt1), Corridor (cologne3, ingolstadt7), and Regional (cologne8, ingolstadt21).

\section{Research Objectives}
\label{sec:Introduction_ResearchObjectives}

The work reported in this thesis pursues five interconnected objectives:

\begin{enumerate}
  \item \textbf{Survey existing TSC approaches.} Review classical, RL-based, and MARL-based methods to identify the state of the art and the limitations that motivate graph-aware, entropy-regularized control.

  \item \textbf{Study local feature representation.} Investigate the effect of local encoder design — specifically MLP encoders versus the FRAP phase-competition encoder — on agent performance and learning dynamics.

  \item \textbf{Study graph-based inter-agent communication.} Analyse whether GATv2-based message passing between neighboring intersections improves coordination, and whether the benefit depends on network size and topology.

  \item \textbf{Study RL optimization schemes.} Compare discrete-action SAC, DQN, and PPO as RL algorithms within the same MARL-TSC framework, examining sample efficiency, stability, and final performance.

  \item \textbf{Evaluate across German RESCO scenarios.} Assess FGS, FGSv2, and FGSv3 pipeline instantiations against classical and deep RL baselines on the German RESCO benchmark, characterizing performance, scalability, and the contribution of individual architectural choices. FGSv2 and FGSv3 emerged iteratively: each generation targets a specific interface failure identified in evaluation of the preceding one, providing both improved performance and empirically grounded design guidelines.
\end{enumerate}

\section{Research Questions and Hypotheses}
\label{sec:Introduction_ResearchQuestions}

The empirical investigation is organized around one overarching main research question and three focused sub-questions that operationalize it.

\medskip
\noindent\textbf{Main Research Question (MRQ):}
\begin{quote}
  \textit{Can a modular three-module MARL pipeline — combining a local encoder, graph-attention communication, and an RL algorithm as independently configurable choices — provide competitive and scalable traffic signal control on real-world German road networks, and which design dimensions are the primary determinants of performance across network scales from single intersections to 21-agent regional clusters?}
\end{quote}

\medskip
The MRQ is operationalized through three sub-questions. Each sub-question addresses a distinct epistemic dimension: external validity (SQ1), internal design analysis (SQ2), and boundary conditions across scale (SQ3).

\medskip
\noindent\textbf{SQ1} \textit{(Baseline comparison):} \textit{How do FGS pipeline instantiations compare in performance with classical traffic-control methods and deep RL baselines across the three German RESCO scale tiers?}

\medskip
\noindent\textbf{SQ2} \textit{(Design-space analysis):} \textit{How do encoder choice (MLP vs.\ FRAP) and graph-attention mechanism (GAT vs.\ GATv2) affect traffic signal control performance and training stability within the FGS pipeline?}

\medskip
\noindent\textbf{SQ3} \textit{(Scale and topology dependency):} \textit{How does the benefit of each design component vary with network scale and topology, from Single-intersection to Regional ($N=21$) scenarios?}

\medskip

Four hypotheses guide the experimental design; they are stated as testable propositions, not guarantees. H1 and H4 inform SQ3; H2 and H3 inform SQ2; all four bear collectively on the MRQ.

\medskip
\noindent\textbf{H1:} \textit{Graph-aware communication improves coordination between neighboring intersections compared with independent-agent methods, as reflected in lower network-level mean travel time and waiting time.}

\medskip
\noindent\textbf{H2:} \textit{Entropy-regularized SAC produces more stable training curves than purely value-based methods (DQN) in selected multi-agent German RESCO scenarios.} This hypothesis is framed as an investigation: SAC may not outperform DQN in all scenarios, and the conditions under which entropy regularization is beneficial are themselves of interest.

\medskip
\noindent\textbf{H3:} \textit{Phase-aware local encoders such as FRAP do not necessarily improve performance in all German RESCO scenarios.} FRAP's phase-competition structure encodes assumptions about movement representation and phase pairing that may not align uniformly with every network topology, making its relative contribution an empirical rather than a guaranteed finding.

\medskip
\noindent\textbf{H4:} \textit{Larger and more complex networks reveal stronger performance differences between independent-agent and graph-aware methods than small single-intersection scenarios.}

\section{Scope and Limitations}
\label{sec:Introduction_ScopeAndLimitations}

\subsection{Scope}
\label{sec:Introduction_Scope}

This thesis focuses on the design, implementation, and empirical evaluation of a graph-aware MARL framework for signalized intersection control. The scope encompasses:
\begin{itemize}
  \item Control algorithm design: the FGS, FGSv2, and FGSv3 architectural variants described in Chapter~\ref{ch:Methodology}.
  \item Algorithm analysis: ablation across encoder types, communication mechanisms, and RL algorithms.
  \item Empirical comparison: six German RESCO scenarios across three scale tiers — \textbf{Single-intersection} (cologne1, ingolstadt1), \textbf{Corridor} (cologne3, ingolstadt7), and \textbf{Regional} (cologne8, ingolstadt21) — against a standardized set of classical and deep RL baselines. This tier classification is used consistently throughout the thesis. FGSv3 was not evaluated on Corridor-tier scenarios; first-generation $\text{FGS}_\text{P}$ results are available for ingolstadt7 and included in the comparison tables.
\end{itemize}

\subsection{Limitations}
\label{sec:Introduction_Limitations}

The following are explicitly outside the scope of this work:

\begin{itemize}
  \item \textbf{Pedestrian and cyclist modeling.} The SUMO environments used do not simulate pedestrian or cyclist movements; the control policy addresses vehicle phase selection only.

  \item \textbf{Simulation environment.} All experiments are conducted in SUMO \citep{lopez2018microscopic} via the TraCI interface. Real-world deployment, with its sensor noise, communication latency, and physical actuator constraints, is not addressed.

  \item \textbf{Scenario construction.} Traffic network files, demand routes, and signal programmes are taken from the RESCO benchmark release without modification. No new traffic scenarios are constructed for this thesis.

  \item \textbf{RL library development.} The thesis does not implement an RL framework from scratch. The implementation builds on RLlib for distributed training, PyTorch Geometric for graph operations, and the SUMO-RL library \citep{alegre2021sumorl} for environment wrapping, each extended with custom modules where required. Earlier exploratory frameworks considered during development include the original RESCO evaluation scripts, CityFlow, LibSignal, and PyTSC; these were not adopted in the final implementation.

  \item \textbf{Reward design.} The reward function is fixed as the differential waiting time (Section~\ref{sec:Methodology_DiffWaitingTime}). Systematic reward shaping or fairness-constrained reward design is not investigated.

  \item \textbf{Continuous action spaces.} The framework operates in discrete phase-selection action spaces throughout; continuous duration assignment is not explored.
\end{itemize}

\section{Thesis Structure}
\label{sec:Introduction_ThesisStructure}

The remainder of this thesis is organized as follows.

\textbf{Chapter~\ref{ch:LiteratureReview}} surveys existing TSC approaches from fixed-time control through single-agent and multi-agent RL, reviews graph neural network variants relevant to inter-agent communication, and identifies the research gap that this thesis addresses.

\textbf{Chapter~\ref{ch:TheoreticalBackground}} establishes the mathematical and conceptual foundations: Markov Decision Processes, partially observable and multi-agent extensions, the progression from dynamic programming to deep RL and entropy-regularized actor-critic methods, graph neural network message passing, and the traffic engineering concepts required to interpret the experiments.

\textbf{Chapter~\ref{ch:Methodology}} describes the FGS framework in detail: observation space, action space, graph construction, baseline algorithms, reward function, and the full architectural specifications of both proposed variants.

\textbf{Chapter~\ref{ch:ExperimentalDesign}} specifies the experimental setup: the TapasCologne and InTAS demand models, SUMO simulation parameters, evaluation metrics and protocols, training hyperparameters, and the end-to-end software pipeline.

\textbf{Chapter~\ref{ch:ResultsAndDiscussion}} presents and interprets the experimental results: main performance comparisons, ablation across architectural variants, cross-scenario analysis, and answers to the three sub-questions and the main research question.

\textbf{Chapter~\ref{ch:ConclusionFutureWork}} summarizes the contributions, reflects on the hypotheses in light of the experimental findings, and identifies directions for future work.
