\chapter{Conclusion and Future Work}
\label{ch:ConclusionFutureWork}

\section{Summary}
\label{sec:ConclusionFutureWork_Summary}

This thesis set out to systematically characterize the three-module design space of MARL-TSC architectures — encoder, graph communication, and RL algorithm — on real-world German road networks, and to derive empirically grounded conclusions about which component choices matter and under what conditions. The investigation was framed not as a claim that any single configuration solves traffic optimization, but as a principled ablation study of how these components interact across a standardized set of scenarios spanning three scale tiers.

Urban traffic signal control presents a multi-agent decision problem with partial observability, heterogeneous intersection topologies, and demand conditions that vary continuously over a simulation episode. Classical methods such as Webster's fixed-time plans and max-pressure control \citep{varaiya2013max} offer tractable, interpretable solutions but cannot adapt dynamically to non-stationary demand. Deep RL approaches such as IntelliLight \citep{wei2018intellilight} and PressLight \citep{wei2019presslight} demonstrated that learned policies can match or exceed adaptive controllers in simulation, but their evaluation was largely confined to synthetic Chinese grid networks. Graph-based MARL methods, most notably CoLight \citep{wei2019colight}, extended RL-based TSC to network-level coordination by treating intersections as nodes in a communication graph. However, none of the published graph-aware MARL methods evaluated on German road networks, none combined a phase-competition-aware local encoder with graph attention, and none used discrete-action SAC as the RL optimiser.

The FGS (\textbf{F}ramework for \textbf{G}raph-aware \textbf{S}ignal control) proposed in this thesis addresses all three gaps through a modular three-module pipeline and a three-generation implementation trajectory. The Dec-POMDP problem formulation (Chapter~\ref{ch:Methodology}) defines each signalized intersection as an agent whose observation is a graph-structured tensor containing local phase state, per-lane density and queue features, and a topological neighborhood constructed via Dijkstra super-edges from the SUMO network file. Actions are discrete phase selections; the reward is the differential waiting time, which provides a dense training signal without requiring domain-specific reward engineering. Three successive FGS instantiations explore the design space through an ablation-driven diagnosis-and-repair cycle. FGSv1 establishes the baseline pipeline, supporting four variants (Table~\ref{tab:fgs_variants}) across two encoder choices (MLP, FRAP) and two communication modes (none, GATv2), and exposes two interface failures. FGSv2 fixes the encoder–communication boundary by preserving per-phase action tokens through the graph aggregation step. FGSv3 fixes the communication signal (switching from competition-derived to demand-aware embeddings) and resolves the critic scalability barrier by replacing the centralized critic with a factored neighborhood critic whose input dimension is $O(d_c)$ independent of $N$, enabling SAC training to complete at $N=21$.

The experimental evaluation (Chapters~\ref{ch:ExperimentalDesign} and~\ref{ch:ResultsAndDiscussion}) assessed FGS, FGSv2, and FGSv3 against seven baselines — fixed-time control, max-pressure, IDQN, IPPO, independent SAC, FRAP, and CoLight — across the six German RESCO scenarios derived from TapasCologne \citep{tapascologne} and InTAS \citep{intasscenario}, organized into the three scale tiers introduced in Section~\ref{sec:Introduction_Scope}.

The experimental findings allow the four hypotheses stated in Chapter~\ref{ch:Introduction} to be revisited in light of evidence:

\textbf{H1} — that graph-aware communication improves coordination compared with independent agents — receives \textit{partial support}. FGS(MLP+GATv2) variants generally reduce network-level mean travel time and mean waiting time relative to their non-graph counterparts in Regional-tier scenarios (cologne8, ingolstadt21). The improvement is most pronounced on ingolstadt21, whose irregular topology provides heterogeneous neighborhood contexts for GATv2 to exploit, and is negligible on Single-intersection and Corridor scenarios where neighborhood information is limited.

\textbf{H2} — that entropy-regularized SAC yields more stable training than DQN — is confirmed in selected scenarios, particularly those with larger intersection counts where the policy landscape is more complex. However, in small single-intersection scenarios (cologne1, ingolstadt1), DQN's simpler update rule converges with comparable or faster stability, consistent with the expectation that entropy bonuses are most valuable where policy diversity matters.

\textbf{H3} — that FRAP-based encoders do not uniformly outperform MLP encoders — is confirmed across all evaluated scenarios. The structural explanation is provided in Section~\ref{sec:ResultsAndDiscussion_Interpretation}: FRAP's phase-competition priors align with regular, four-directional intersection layouts but produce misaligned features on German RESCO networks with heterogeneous phase counts. This finding does not diminish FRAP's theoretical motivation; it highlights that the benefit of structured local encoding depends on the match between the encoder's priors and the target network topology.

\textbf{H4} — that Regional-tier networks reveal stronger differences between independent and graph-aware methods than Single-intersection or Corridor scenarios — is confirmed in the direction expected: the performance gap between FGS(MLP) and FGS(MLP+GATv2) grows with the tier scale. Single-intersection scenarios (cologne1, ingolstadt1) show negligible graph-communication benefit by definition; Regional-tier scenarios (cologne8, ingolstadt21) show the clearest advantage for GATv2 variants, where upstream spillback propagation provides a meaningful signal for neighborhood aggregation.

Taken together, the results demonstrate that the FGS pipeline provides a productive framework for characterizing MARL-TSC design on German road networks. The three-generation trajectory — driven by empirical findings rather than a priori design — confirms that encoder choice, graph-attention expressiveness, and critic architecture each govern a distinct failure mode, and that addressing all three is necessary for competitive performance at Regional scale. FGSv3-PPO achieves 22.7~s mean waiting time on ingolstadt21, outperforming all seven baselines including CoLight ($-81\%$) and standalone PPO ($-12\%$), while the ablation establishes concrete design guidelines: MLP encoders outperform phase-competition encoders on German intersection heterogeneity; GATv2 provides a 66\% reduction over GAT at $N=21$; and SAC requires a factored critic to complete training at Regional scale.

\section{Contributions}
\label{sec:ConclusionFutureWork_Contributions}

This thesis makes five contributions at different levels of abstraction.

\subsection{Framework Contribution}
\label{sec:ConclusionFutureWork_FrameworkContribution}

A modular, reproducible experimental stack for graph-aware MARL traffic signal control was designed and implemented. The stack integrates SUMO \citep{lopez2018microscopic} via the TraCI interface, the SUMO-RL gymnasium wrapper \citep{alegre2021sumorl}, RLlib for distributed multi-agent training, PyTorch Geometric for graph neural network operations, Hydra for structured configuration management, and Weights \& Biases \citep{wandb} for experiment tracking. Graph-structured observations are constructed automatically from the SUMO network file, and the training pipeline supports arbitrary agent count, flexible phase structures, and configurable encoder and communication modules. The implementation is publicly available and extends the open-source SUMO-RL library with graph observation wrappers, custom RLlib policy modules, and reproducible experiment scripts.

\subsection{Algorithmic Contribution}
\label{sec:ConclusionFutureWork_AlgorithmicContribution}

Three successive FGS pipeline instantiations were designed and implemented (Chapter~\ref{ch:Methodology}), each addressing a specific module-interface failure identified in evaluation of the preceding generation. FGSv1 (FGS) establishes the baseline, exposing that FRAP's scalar projection discards per-phase information before graph aggregation and that the centralized critic's $O(Nd)$ input prevents SAC from completing training at $N=21$. FGSv2 fixes the Module-I→Module-II boundary by tapping FRAP before scalar projection, preserving per-phase action tokens for the graph layer; this exposes that the communication signal carries competition-derived rankings rather than demand state. FGSv3 fixes both remaining failures: the Module-II signal is replaced with demand-aware embeddings, and the Module-III centralized critic is replaced with a factored neighborhood critic ($O(d_c + A_{\max})$ independent of $N$), enabling both SAC and PPO to complete training at $N=21$. Full architectural details are given in Sections~\ref{sec:Methodology_FGSv2}--\ref{sec:Methodology_FGSv3}.

\subsection{Empirical Contribution}
\label{sec:ConclusionFutureWork_EmpiricalContribution}

A comprehensive evaluation comparing FGS, FGSv2, and FGSv3 against seven baseline methods — fixed-time control, max-pressure, IDQN, IPPO, independent SAC, FRAP, and CoLight — across all six German RESCO scenarios was conducted. The evaluation protocol targets the RESCO benchmark standard of 200 training episodes, evaluation every five episodes, and performance reported on mean travel time, mean delay, mean waiting time, and throughput. Due to resource constraints, most configurations correspond to single-seed runs (seed~0); multi-seed averaging over five seeds with standard-deviation reporting remains as a priority for follow-on validation (Section~\ref{sec:ResultsAndDiscussion_Limitations}).

\subsection{Analytical Contribution}
\label{sec:ConclusionFutureWork_AnalyticalContribution}

A component-level ablation study disentangles the contribution of each architectural element: encoder type (MLP vs.\ FRAP), communication (with vs.\ without GATv2), and RL algorithm (SAC vs.\ DQN vs.\ PPO). This analysis quantifies the conditions under which phase-aware encoding helps or hurts, and the network scales at which graph communication produces the most consistent benefit. The finding that FRAP's inductive bias does not consistently transfer to German intersection topologies is itself an empirical contribution to the broader MARL-TSC literature, where FRAP has been widely adopted without topology-specific validation.

\subsection{Conceptual Contribution}
\label{sec:ConclusionFutureWork_ConceptualContribution}

The thesis situates decentralized multi-agent TSC within the Organic Computing paradigm: each intersection controller operates on local and neighborhood-level information, with no centralized coordinator at execution time, yet the system produces coherent network-level traffic behavior from local interactions. This framing provides a principled lens for interpreting the empirical results — in particular, the question of whether agent coordination emerges from graph-based communication or from independent reward optimization alone — and connects the engineering contribution to a broader tradition of self-organizing multi-agent systems.
% TODO: add a proper Organic Computing citation here (e.g.\ M{\"u}ller-Schloer 2004 or W{\"u}rtz 2008); \cite{lowe2017multi} (MADDPG) was incorrectly placed here previously.

\section{Future Research Directions}
\label{sec:ConclusionFutureWork_FutureResearch}

The work reported in this thesis opens several directions for further investigation.

\paragraph{Continuous green-time optimization.}
The FGS framework selects discrete phases; phase duration is fixed by the minimum green time constraint. Extending the action space to include duration assignment — either as a continuous output or as a discrete time-slot selection — would allow the controller to optimize both phase sequence and green-time allocation jointly. Parameterized action spaces or hybrid discrete-continuous actors are natural candidates for this extension.

\paragraph{Fairness-constrained reward design.}
The differential waiting time reward aggregates delay across all lanes without distinguishing between high-demand and low-demand directions. Future work could incorporate explicit fairness constraints — for example, penalizing extreme queue disparities across movements within an intersection or across intersections in the network — to produce more equitable service distributions without sacrificing overall throughput.

\paragraph{Transfer learning and domain adaptation.}
The FGS models trained on TapasCologne and InTAS demand profiles were evaluated on the same demand distribution used during training. An important open question is whether policies generalize across German cities, to demand patterns not represented in the training data, or across different countries' intersection geometries. Meta-learning or domain-randomization approaches could be used to train controllers robust to demand distribution shift.

\paragraph{Real-world deployment.}
All experiments in this thesis were conducted in SUMO simulation. Deployment to real intersection controllers introduces additional challenges: sensor noise from inductive loop detectors, communication latency between the central coordination layer and field controllers, legal constraints on minimum phase durations, and the absence of a differentiable environment for online policy gradient updates. Sim-to-real transfer methods and hardware-in-the-loop evaluation are natural next steps.

\paragraph{Scalability beyond the Regional tier.}
The largest evaluated scenario (ingolstadt21, Regional tier, $N=21$) contains twenty-one intersections. The FGSv3 factored neighborhood critic reduces critic input dimensionality from $O(Nd_{h'})$ to $O(d_c)$, enabling SAC training to complete at $N=21$ where FGSv2 failed. However, the GATv2 message-passing layer itself has $O(N)$ memory in the number of nodes, and the graph construction from the SUMO network file currently targets small-to-moderate scenarios. Extending FGS beyond the Regional tier to hundreds of intersections will likely require hierarchical MARL architectures — in which local clusters of intersections are managed by sub-controllers that themselves communicate through a higher-level graph — or sparse graph attention variants that limit each agent's receptive field to a fixed-size neighborhood regardless of network size.

\paragraph{Dynamic graph topology.}
The graph structure used in FGS is constructed offline from the SUMO network file and remains fixed throughout the episode. Real traffic networks experience topology changes due to incidents, road closures, and temporary signal failures. A dynamic graph construction mechanism that updates edge weights and neighborhood sets in response to real-time travel-time estimates would make the framework more robust to network perturbations.

\paragraph{Graph attention interpretability.}
FGSv2's GATv2 attention layer assigns learned weights to each neighbor, potentially encoding which upstream or downstream intersections are most relevant under current traffic conditions. Visualizing attention weights over time and correlating them with queue lengths and phase states could provide insight into the implicit coordination strategies learned by the controller, and connect the results more concretely to the Organic Computing interpretation of self-organized traffic management.

\paragraph{Pedestrian and multi-modal integration.}
The current formulation addresses vehicle phase selection only. Extending the observation space and reward function to account for pedestrian crossing demands, cycle lanes, and public transport priority would substantially increase the realism and practical applicability of the framework in European urban environments, where multi-modal integration is a primary concern of traffic management authorities.
