\chapter{Tools and Technologies}
\label{ch:Appendix_ToolsAndTechnologies}

The FGS experimental pipeline integrates the following software components.
\textbf{SUMO 1.18} \citep{lopez2018microscopic} provides microscopic traffic simulation and exposes the TraCI socket interface for per-step observation and action exchange.
\textbf{SUMO-RL} \citep{alegre2021sumorl} wraps SUMO environments as Gymnasium-compatible multi-agent environments; graph observation wrappers and custom reward functions were added for this work.
\textbf{RLlib 2.x} (Ray) provides the distributed multi-agent training loop; custom policy modules subclassing \texttt{TorchRLModule} implement the FGS actor, critic, and encoder components.
\textbf{PyTorch Geometric} implements the GATv2 convolution (\texttt{GATv2Conv}) and heterogeneous graph batching utilities.
\textbf{Hydra} structures all experiment configurations as composable YAML files; scenario, algorithm, and architecture parameters are varied without code changes.
\textbf{Weights \& Biases} \citep{wandb} tracks all training runs, logs per-episode RESCO metrics, and stores model checkpoints.

\chapter{Mathematical Objectives, Gradients, and Stability Analysis}
\label{ch:Appendix_LossesObjectives}

This appendix collects the mathematical derivations that underpin the FGS/FGSv2 framework: objective functions, loss gradients, masked action distributions, architectural equivariance properties, and the theoretical basis for convergence claims. Proofs are stated for the finite tabular/exact-policy setting; Section~\ref{sec:Appendix_convergence_disclaimer} states explicitly which guarantees do not extend to the implemented deep MARL system.

% ─────────────────────────────────────────────────────────────────────────────
\section{Notation and CTDE Objective}
\label{sec:Appendix_CTDE}
% ─────────────────────────────────────────────────────────────────────────────

Throughout this appendix: $N$ is the number of agents; $A_i = |\mathcal{A}_i|$ the valid phase count for agent $i$; $A_{\max} = \max_i A_i$ the padded action-space width; $o_i^t$ the local observation of agent $i$ at step $t$; $\mathbf{o}_t = (o_1^t,\ldots,o_N^t)$ the joint observation; $\mathbf{a}_t = (a_1^t,\ldots,a_N^t)$ the joint action.

The cooperative multi-agent objective averages individual rewards:
\begin{equation}
  J(\phi) = \mathbb{E}_{\pi_\phi}\!\left[\sum_{t=0}^{T}\gamma^t \frac{1}{N}\sum_{i=1}^{N}r_i^t\right].
  \label{eq:marl_objective}
\end{equation}
Under maximum-entropy augmentation this becomes
\begin{equation}
  J_{\mathrm{MaxEnt}}(\phi) = \mathbb{E}_{\pi_\phi}\!\left[\sum_{t=0}^{T}\gamma^t \frac{1}{N}\sum_{i=1}^{N}\Bigl(r_i^t + \alpha\,\mathcal{H}\!\bigl(\pi_\phi(\cdot\mid o_i^t)\bigr)\Bigr)\right].
  \label{eq:marl_maxent}
\end{equation}
Agents optimize per-agent SAC objectives with local rewards $r_i^t$; evaluation follows the RESCO network-level average protocol. The CTDE factorization is
\begin{equation}
  \pi_\phi(\mathbf{a}\mid\mathbf{o}) = \prod_{i=1}^{N}\pi_\phi(a_i\mid o_i, \mathcal{G}),
\end{equation}
while at execution time each agent $i$ acts on its local embedding $\mathbf{h}_i'$ alone.

% ─────────────────────────────────────────────────────────────────────────────
\section{Differential Waiting-Time Reward: Telescoping Property}
\label{sec:Appendix_reward_telescope}
% ─────────────────────────────────────────────────────────────────────────────

The differential waiting-time reward is $r_t = W_{t-1} - W_t$ (Equation~\ref{eq:reward}). The following result shows that maximizing the discounted sum of differential rewards is equivalent to minimizing a discounted sum of cumulative waiting times.

\begin{proposition}[Reward telescoping]
\label{prop:telescope}
For a finite horizon $T$ and discount factor $\gamma\in(0,1)$,
\begin{equation}
  \sum_{t=1}^{T}\gamma^{t-1}(W_{t-1}-W_t)
  = W_0
    - (1-\gamma)\sum_{t=1}^{T-1}\gamma^{t-1}W_t
    - \gamma^{T-1}W_T.
  \label{eq:telescope}
\end{equation}
\end{proposition}

\begin{proof}
Separate the sum: $S = \sum_{t=1}^{T}\gamma^{t-1}W_{t-1} - \sum_{t=1}^{T}\gamma^{t-1}W_t$.
Re-index the first sum ($s \leftarrow t-1$, range $0$ to $T-1$):
\begin{align*}
  S &= \sum_{t=0}^{T-1}\gamma^{t}W_t
       - \sum_{t=1}^{T}\gamma^{t-1}W_t \\
    &= W_0
       + \sum_{t=1}^{T-1}\gamma^{t}W_t
       - \frac{1}{\gamma}\sum_{t=1}^{T-1}\gamma^{t}W_t
       - \gamma^{T-1}W_T \\
    &= W_0
       + \!\Bigl(1-\tfrac{1}{\gamma}\Bigr)\!\sum_{t=1}^{T-1}\gamma^{t}W_t
       - \gamma^{T-1}W_T \\
    &= W_0
       - (1-\gamma)\sum_{t=1}^{T-1}\gamma^{t-1}W_t
       - \gamma^{T-1}W_T.
\end{align*}
\end{proof}

\paragraph{Interpretation.} Maximizing $S$ over trajectories is equivalent (up to the constant $W_0$) to minimizing $(1-\gamma)\sum_{t=1}^{T-1}\gamma^{t-1}W_t + \gamma^{T-1}W_T$: a discounted sum of intermediate waiting times plus the terminal waiting time. When $\gamma<1$ both terms contribute; as $\gamma\to 1$ the sum approaches $W_0 - W_T$, so the agent would primarily optimize terminal rather than cumulative waiting. Using $\gamma=0.99$ in this work keeps the intermediate terms substantial.

% ─────────────────────────────────────────────────────────────────────────────
\section{Masked Discrete Action Distribution}
\label{sec:Appendix_masked_softmax}
% ─────────────────────────────────────────────────────────────────────────────

Intersections in the German RESCO scenarios have heterogeneous phase counts; invalid phases must receive zero probability. Define the valid phase set $\mathcal{A}_i(o_i) = \{a : m_{i,a}=1\}$ and the masked policy
\begin{equation}
  \pi_\phi(a\mid o_i)
  = \frac{m_{i,a}\exp z_{i,a}}
         {\displaystyle\sum_{b=1}^{A_{\max}}m_{i,b}\exp z_{i,b}}.
  \label{eq:masked_softmax}
\end{equation}

\begin{proposition}[Zero probability and zero gradient on masked actions]
\label{prop:masked_zero}
If $m_{i,a}=0$ then $\pi_\phi(a\mid o_i)=0$ and $\partial\pi_\phi(a\mid o_i)/\partial z_{i,b}=0$ for all $b$.
\end{proposition}
\begin{proof}
When $m_{i,a}=0$ the numerator of~\eqref{eq:masked_softmax} is zero regardless of $z_{i,a}$, so $\pi_\phi(a\mid o_i)=0$. Since this holds for all values of all logits, every partial derivative is also zero.
\end{proof}

\paragraph{Implementation.} In practice invalid logits are set to $-10^9$ before the standard softmax rather than using~\eqref{eq:masked_softmax} directly; the two formulations are numerically equivalent when the finite approximation dominates all valid logits.

% ─────────────────────────────────────────────────────────────────────────────
\section{Discrete SAC: Losses and Gradients}
\label{sec:Appendix_SAC_gradients}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Critic Loss and Gradient}
\label{sec:Appendix_critic_loss}

Given a replay-buffer sample $(s,a,r,s',d)$, the soft TD target (with stop-gradient) is
\begin{equation}
  y = r + \gamma(1-d)\sum_{a'}\pi_\phi(a'\mid s')\!
      \left[\min_{j=1,2}Q_{\bar\theta_j}(s',a') - \alpha\log\pi_\phi(a'\mid s')\right].
\end{equation}
For twin critic $j$:
\begin{equation}
  \mathcal{L}_Q(\theta_j)
  = \mathbb{E}_{\mathcal{D}}\!\left[\tfrac{1}{2}\bigl(Q_{\theta_j}(s,a)-y\bigr)^2\right],
  \qquad
  \nabla_{\!\theta_j}\mathcal{L}_Q
  = \mathbb{E}_{\mathcal{D}}\!\left[\bigl(Q_{\theta_j}(s,a)-y\bigr)\nabla_{\!\theta_j}Q_{\theta_j}(s,a)\right].
\end{equation}
With Prioritized Experience Replay the loss is reweighted by importance-sampling corrections. For TD error $\delta_k = Q_{\theta_j}(s_k,a_k)-y_k$, define
\begin{equation}
  p_k = |\delta_k|+\varepsilon,\quad
  P(k) = \frac{p_k^\beta}{\sum_j p_j^\beta},\quad
  w_k = \!\left(\frac{1}{|\mathcal{D}|\,P(k)}\right)^{\!\rho},\quad
  \tilde w_k = \frac{w_k}{\max_j w_j},
  \label{eq:per_weights}
\end{equation}
giving PER critic loss $\mathcal{L}_Q(\theta_j) = \mathbb{E}_{k\sim P}[\tilde w_k\,\tfrac{1}{2}\delta_k^2]$.

\subsection{Actor Loss and Softmax-Logit Gradient}
\label{sec:Appendix_actor_gradient}

The discrete actor loss is (reproduced from Equation~\ref{eq:sac_actor_loss}):
\begin{equation}
  \mathcal{L}_\pi(\phi) = \mathbb{E}_{s\sim\mathcal{D}}\!\left[
    \sum_{a\in\mathcal{A}(s)}\pi_\phi(a\mid s)
    \Bigl(\alpha\log\pi_\phi(a\mid s) - \min_k Q_{\theta_k}(s,a)\Bigr)\right].
\end{equation}
Define the soft state value $V_\theta^\pi(s) = \sum_a\pi_a\bigl[Q_\theta(s,a)-\alpha\log\pi_a\bigr]$.

\begin{proposition}[Logit gradient of the actor loss]
\label{prop:actor_gradient}
Let $\pi_a = \operatorname{softmax}(z)_a$. Then
\begin{equation}
  \frac{\partial\mathcal{L}_\pi}{\partial z_b}
  = \pi_b\Bigl[\alpha\log\pi_b - Q_\theta(s,b) + V_\theta^\pi(s)\Bigr].
  \label{eq:actor_logit_grad}
\end{equation}
\end{proposition}
\begin{proof}
Using the softmax Jacobian $\partial\pi_a/\partial z_b = \pi_a(\delta_{ab}-\pi_b)$ and applying the product rule:
\begin{align*}
  \frac{\partial\mathcal{L}_\pi}{\partial z_b}
  &= \sum_a\frac{\partial\pi_a}{\partial z_b}
     \bigl(\alpha\log\pi_a - Q_\theta(s,a) + \alpha\bigr) \\
  &= \sum_a\pi_a(\delta_{ab}-\pi_b)
     \bigl(\alpha\log\pi_a - Q_\theta(s,a) + \alpha\bigr) \\
  &= \pi_b\bigl(\alpha\log\pi_b - Q_\theta(s,b) + \alpha\bigr)
     - \pi_b\!\sum_a\pi_a\bigl(\alpha\log\pi_a - Q_\theta(s,a) + \alpha\bigr).
\end{align*}
The inner sum equals $-V_\theta^\pi(s)+\alpha$. Substituting and cancelling $\alpha$ yields~\eqref{eq:actor_logit_grad}.
\end{proof}

\paragraph{Interpretation.} Actions with $Q_\theta(s,b) > V_\theta^\pi(s)$ receive a negative gradient component and therefore increased probability; the baseline $V_\theta^\pi(s)$ centers each update; the $\alpha\log\pi_b$ term moderates excessive concentration on any single action.

\subsection{Temperature Objective and Log-Parameterization}
\label{sec:Appendix_alpha_loss}

The temperature $\alpha$ is updated by minimizing (Equation~\ref{eq:sac_alpha_loss})
\begin{equation}
  \mathcal{L}(\alpha)
  = \mathbb{E}_{s\sim\mathcal{D}}\!\left[
    \alpha\bigl(\mathcal{H}(\pi(\cdot\mid s)) - \mathcal{H}_{\mathrm{target}}\bigr)\right].
  \label{eq:alpha_loss_appendix}
\end{equation}
Gradient $\nabla_\alpha\mathcal{L} = \mathbb{E}[\mathcal{H}(\pi)-\mathcal{H}_{\mathrm{target}}]$ correctly drives $\alpha$ upward when entropy falls below the target and downward when it exceeds it.

For numerical stability the implementation optimizes $\beta = \log\alpha$ (ensuring $\alpha=e^\beta>0$):
\begin{equation}
  \mathcal{L}_\alpha(\beta)
  = e^\beta\,\mathbb{E}_{s\sim\mathcal{D}}\!\left[
    \mathcal{H}_{\mathrm{target}} - \mathcal{H}(\pi_\phi(\cdot\mid s))\right],
  \qquad
  \nabla_\beta\mathcal{L}_\alpha
  = e^\beta\,\mathbb{E}\!\left[\mathcal{H}_{\mathrm{target}} - \mathcal{H}(\pi)\right].
\end{equation}
This is the sign-flipped reparameterization of~\eqref{eq:alpha_loss_appendix}; both forms drive $\alpha$ in the same direction under their respective gradient-descent steps.

% ─────────────────────────────────────────────────────────────────────────────
\section{PPO: Generalized Advantage Estimation and Joint Loss}
\label{sec:Appendix_PPO}
% ─────────────────────────────────────────────────────────────────────────────

The TD residual and GAE advantage estimate are
\begin{align}
  \delta_t &= r_t + \gamma V_\psi(s_{t+1}) - V_\psi(s_t), \label{eq:td_residual} \\
  \hat{A}_t^{\mathrm{GAE}} &= \sum_{l=0}^{T-t-1}(\gamma\lambda)^l\delta_{t+l},
  \label{eq:gae}
\end{align}
with $\lambda=0.95$ and $\gamma=0.99$; value targets are $\hat{R}_t = \hat{A}_t + V_\psi(s_t)$.

The joint PPO objective (to maximize) is
\begin{equation}
  \mathcal{L}_{\mathrm{PPO}} = \mathcal{L}^{\mathrm{CLIP}} - c_1\mathcal{L}^{VF} + c_2\mathcal{H}[\pi_\phi],
  \quad c_1=0.5,\quad c_2=10^{-3},
  \label{eq:ppo_joint}
\end{equation}
where $\mathcal{L}^{\mathrm{CLIP}}$ is defined in Equation~\ref{sec:Methodology_PPO} and $\mathcal{L}^{VF} = \mathbb{E}_t[(V_\psi(s_t)-\hat{R}_t)^2]$.

% ─────────────────────────────────────────────────────────────────────────────
\section{GATv2: Edge Weights and Permutation Equivariance}
\label{sec:Appendix_GATv2_properties}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Edge-Weight Integration}
\label{sec:Appendix_GATv2_edge_weights}

The graph construction assigns $w_{ij}=1/\tau_{ij}$ (Equation~\ref{eq:edge_weight}). If incorporated into GATv2, the unnormalized attention score can be augmented as
\begin{equation}
  e_{ij} = \mathbf{a}^\top\mathrm{LeakyReLU}\!\bigl(\mathbf{W}[\mathbf{h}_i\,\Vert\,\mathbf{h}_j]\bigr)
           + \eta\log(w_{ij}+\varepsilon),
  \label{eq:gatv2_edge_weight}
\end{equation}
where $\eta>0$ is a learnable scalar and $\varepsilon>0$ prevents $\log 0$. In the current implementation $w_{ij}$ is passed to \texttt{GATv2Conv} as edge-attribute metadata but the log-weight additive term in~\eqref{eq:gatv2_edge_weight} is not activated; topology enters only through graph connectivity. Equation~\eqref{eq:gatv2_edge_weight} is presented as a candidate extension.

\subsection{Node-Permutation Equivariance}
\label{sec:Appendix_GATv2_equivariance}

Let $P\in\{0,1\}^{N\times N}$ be a node-permutation matrix, $X\in\mathbb{R}^{N\times d}$ the feature matrix, and $\mathcal{E}$ the edge set.

\begin{proposition}[GATv2 node-permutation equivariance]
\label{prop:gatv2_equivariance}
The GATv2 encoder $f$ satisfies $f(PX, P\mathcal{E}) = Pf(X,\mathcal{E})$.
\end{proposition}
\begin{proof}
\begin{enumerate}
\item \emph{Shared node encoder.} The local encoder $f_{\mathrm{enc}}$ uses the same weights for every node; permuting input rows gives $(f_{\mathrm{enc}}(PX))_i = f_{\mathrm{enc}}(X_{P^{-1}i}) = (Pf_{\mathrm{enc}}(X))_i$.
\item \emph{Attention coefficients.} Under relabelling $i\mapsto Pi$, the logit $e_{Pi,Pj}$ is evaluated at features $(h_{Pi},h_{Pj})$, equalling $e_{ij}$ computed from the permuted embeddings. Softmax normalization over neighborhood $\{Pk : k\in\mathcal{N}(i)\}$ is identical.
\item \emph{Neighborhood aggregation.} Message pooling sums over unordered neighbor sets; the neighbor set of $Pi$ in $P\mathcal{E}$ is $\{Pj : j\in\mathcal{N}(i)\}$, which matches.
\item \emph{Combined.} Relabelling inputs relabels outputs: $f(PX,P\mathcal{E})_i = f(X,\mathcal{E})_{P^{-1}i}$, i.e.\ $f(PX,P\mathcal{E})=Pf(X,\mathcal{E})$. \qedhere
\end{enumerate}
\end{proof}

\paragraph{Significance.} Equivariance is what makes parameter sharing across intersections geometrically consistent: a policy trained on any intersection generalizes to any topologically similar intersection without retraining.

% ─────────────────────────────────────────────────────────────────────────────
\section{FRAP / FGSv2: Phase-Permutation Equivariance}
\label{sec:Appendix_FRAP_equivariance}
% ─────────────────────────────────────────────────────────────────────────────

Let $P_A$ permute the $A_i$ candidate phases of intersection $i$. The FRAP encoder $g$ satisfies
\begin{equation}
  g\bigl(P_A\mathbf{M}_{\mathrm{pair}},\; P_A\mathbf{M}_{\mathrm{comp}}P_A^\top\bigr)
  = P_A\,g\bigl(\mathbf{M}_{\mathrm{pair}},\mathbf{M}_{\mathrm{comp}}\bigr).
  \label{eq:frap_equivariance}
\end{equation}
This holds because (1) movement embeddings share parameters across movements; (2) phase-pair pooling (Equation~\ref{eq:frap_phase_emb}) applies the same scaled sum to each candidate phase independently; (3) pairwise competition features (Equation~\ref{eq:frap_c}) use the same convolution for every ordered pair $(a,b)$; (4) the FGSv2 mean-pool over competitor set $V_{i,a}$ is permutation-invariant over competitors (Equation~\ref{eq:fgsv2_token}).

\paragraph{Caveat.} FGSv2's action-conditioned actor concatenates a one-hot phase identity $\mathbf{e}_a$ (Equation~\ref{eq:fgsv2_actor}). For~\eqref{eq:frap_equivariance} to hold exactly, this one-hot must transform as $\mathbf{e}_a\mapsto P_A\mathbf{e}_a$ under phase relabelling. Since phases are indexed consistently across all training intersections this does not affect learning, but formal phase-permutation equivariance of FGSv2 holds only up to the phase-identity encoding.

% ─────────────────────────────────────────────────────────────────────────────
\section{Centralized Critic: Input-Dimension Scaling}
\label{sec:Appendix_critic_complexity}
% ─────────────────────────────────────────────────────────────────────────────

For FGSv2 (Equation~\ref{eq:fgsv2_critic}), the critic input concatenates: (i) flattened gated graph contexts ($Nd_{h'}$ dims); (ii) joint action-probability matrix ($NA_{\max}$ dims); (iii) agent $k$'s action token ($d_z$ dims); (iv) agent $k$'s graph context ($d_{h'}$ dims); (v) one-hot agent identity ($N$ dims). The total is
\begin{equation}
  d_{\mathrm{critic}} = N(d_{h'} + A_{\max}) + d_z + d_{h'} + N
                       = O\!\bigl(N(d_{h'} + A_{\max})\bigr).
  \label{eq:critic_dim}
\end{equation}
At ingolstadt21 ($N=21$, $d_{h'}=128$, $A_{\max}=6$): $d_{\mathrm{critic}} \approx 21\times 134 + 277 = 3{,}091$. Twin critics, per-agent evaluation, and a PER replay batch multiply effective memory consumption considerably, which is the proximate cause of the resource exhaustion reported in Section~\ref{sec:ResultsAndDiscussion_Convergence} for the original FGS(MLP+GATv2+SAC) run on ingolstadt21, which terminated after 17 of 200 training episodes. PPO's per-agent value function has input dimension $O(d_{h'})$, independent of $N$, giving it a substantially smaller memory footprint at regional scale.

% ─────────────────────────────────────────────────────────────────────────────
\section{FGSv3: Factored Critic Input-Dimension Bound}
\label{sec:Appendix_fgsv3_critic_complexity}
% ─────────────────────────────────────────────────────────────────────────────

The FGSv3 factored neighborhood critic (Equation~\ref{eq:fgsv3_critic}) replaces the full graph-context concatenation of FGSv2 with a sum-pooled neighborhood aggregate. For agent $k$, the critic input is:
\begin{enumerate}
  \item[(i)] agent $k$'s action token $\mathbf{z}_{k,a}$ ($d_z$ dims);
  \item[(ii)] agent $k$'s gated graph context $\mathbf{g}_k$ ($d_c$ dims);
  \item[(iii)] the neighbor graph-context pool $\mathbf{h}_{\mathrm{nbr},k} = \sum_{j \in \mathcal{N}(k)} \tilde{w}_{kj}\,\mathbf{g}_j$ ($d_c$ dims);
  \item[(iv)] the neighbor action pool $\mathbf{a}_{\mathrm{nbr},k} = \sum_{j \in \mathcal{N}(k)} \tilde{w}_{kj}\,\mathbf{a}_{t-1,j}$ ($A_{\max}$ dims);
  \item[(v)] the candidate-action identity $\mathbf{e}_a$ ($A_{\max}$ dims).
\end{enumerate}

\begin{proposition}[Factored critic input dimension]
\label{prop:fgsv3_critic_dim}
The FGSv3 factored critic input dimension is
\begin{equation}
  d_{\mathrm{critic}}^{v3} = d_z + 2d_c + 2A_{\max} = O\!\bigl(d_c + A_{\max}\bigr),
  \label{eq:fgsv3_critic_dim}
\end{equation}
independent of the number of agents $N$.
\end{proposition}

\begin{proof}
Items (iii) and (iv) above are computed by weighted sum-pooling over $\mathcal{N}(k)$. Since each sum produces a single vector in $\mathbb{R}^{d_c}$ (resp.\ $\mathbb{R}^{A_{\max}}$), the output dimension is independent of $|\mathcal{N}(k)|$ and therefore independent of $N$. Items (i)--(v) contribute $d_z + d_c + d_c + A_{\max} + A_{\max} = d_z + 2d_c + 2A_{\max}$. With $d_z$ and $d_c$ fixed hyperparameters, the expression is $O(d_c + A_{\max})$. \qedhere
\end{proof}

\paragraph{Numerical comparison.}
At ingolstadt21 ($N=21$, $d_z=d_c=64$, $A_{\max}=6$):
\[
  d_{\mathrm{critic}}^{v3} = 64 + 2(64) + 2(6) = 204,
\]
compared to $d_{\mathrm{critic}}^{v2} = 3{,}091$ from Equation~\eqref{eq:critic_dim}. The factored critic reduces the input dimension by a factor of approximately 15. Note that FGSv3 narrows the adapter (action-token) dimension from $d_z=128$ (FGSv2) to $d_z=64$ to match the demand communication dimension $d_c$; this joint choice yields the 204-dimensional input above. Empirically, FGSv2-SAC completes 200 training episodes on ingolstadt21 with a best-validation delay of 53.7~s; FGSv3-SAC achieves 45.3~s — a 16\% improvement — while operating with a 15$\times$ smaller critic input.

\paragraph{Information loss relative to centralized critic.}
The factored critic observes the graph contexts and previous actions of direct neighbors only, not of all agents. Any coordination signal that requires $k$ to be informed about agents more than one hop away is lost. However, FGSv3's GATv2 layer already aggregates two-hop contextual information into $\mathbf{g}_k$ (one hop of message-passing updates $\mathbf{g}_k$ from its direct neighbors, each of which has already aggregated from \emph{their} neighbors via the shared FRAP--GATv2 pipeline). The factored critic therefore has access to two-hop demand information implicitly through $\mathbf{g}_k$, even though it only inspects one-hop neighbor contexts $\{\mathbf{g}_j : j \in \mathcal{N}(k)\}$ explicitly. This argument does not constitute a formal equivalence — the centralized critic can in principle learn coordination functions that the factored critic cannot — but it motivates the factored critic as a practical approximation \citep{lowe2017multi}.

% ─────────────────────────────────────────────────────────────────────────────
\section{FGSv3: Demand Communication Branch Equivariance}
\label{sec:Appendix_fgsv3_equivariance}
% ─────────────────────────────────────────────────────────────────────────────

\begin{proposition}[Demand branch node-permutation equivariance]
\label{prop:fgsv3_node_equivariance}
Let $P$ be a permutation of the $N$ agent indices and let $P\mathcal{E}$ denote the correspondingly relabelled edge set. The FGSv3 demand communication branch $\mathbf{d}: \mathbb{R}^{N \times \ldots} \to \mathbb{R}^{N \times d_c}$ satisfies
\[
  \mathbf{d}(P\mathbf{X}, P\mathcal{E}) = P\,\mathbf{d}(\mathbf{X}, \mathcal{E}),
\]
i.e.\ the demand embeddings are permuted consistently with the node relabelling.
\end{proposition}

\begin{proof}
The demand branch (Equation~\ref{eq:fgsv3_demand}) applies the same computation to each node $i$ independently: $\mathbf{d}_i = \operatorname{LayerNorm}(\operatorname{ReLU}(W_{\mathrm{comm}}[\bar{\boldsymbol{\phi}}_i \,\Vert\, \operatorname{Emb}_{\mathrm{phase}}(a_{t-1,i})]))$. Because the weights $W_{\mathrm{comm}}$, $\operatorname{LayerNorm}$, and $\operatorname{Emb}_{\mathrm{phase}}$ are shared across all nodes, relabelling node $i \mapsto Pi$ yields $\mathbf{d}_{Pi}(\mathbf{X}) = \mathbf{d}_i(\mathbf{X})$ evaluated at the feature row of $Pi$, which equals $\mathbf{d}_i(P\mathbf{X})_{Pi}$. Hence $\mathbf{d}(P\mathbf{X}, P\mathcal{E})_i = \mathbf{d}(X, \mathcal{E})_{P^{-1}i}$, i.e.\ $\mathbf{d}(P\mathbf{X}, P\mathcal{E}) = P\,\mathbf{d}(\mathbf{X}, \mathcal{E})$. \qedhere
\end{proof}

\paragraph{Phase-permutation equivariance.}
The mean-pooling in Equation~\eqref{eq:fgsv3_phi_mean} is invariant under phase permutations $P_A$: $\bar{\boldsymbol{\phi}}_i$ equals the mean of the relabelled phase embeddings, which equals the mean of the original embeddings since the mean is unordered. Therefore the demand branch is phase-permutation \emph{invariant} (the output $\mathbf{d}_i$ does not depend on the phase ordering), subject to the same caveat as FGSv2 (Section~\ref{sec:Appendix_FRAP_equivariance}): the previous-action embedding $\operatorname{Emb}_{\mathrm{phase}}(a_{t-1,i})$ is indexed by phase label, so formal equivariance under phase relabelling requires the embedding table to transform consistently with $P_A$.

\paragraph{Node-permutation equivariance of the full FGSv3 pipeline.}
Combining Proposition~\ref{prop:fgsv3_node_equivariance} with the existing results for FRAP action tokens (Section~\ref{sec:Appendix_FRAP_equivariance}) and GATv2 (Section~\ref{sec:Appendix_GATv2_properties}): (i) the demand branch is node-equivariant; (ii) GATv2 applied to $\{\mathbf{d}_i\}$ with the permuted edge set $P\mathcal{E}$ is node-equivariant (by the GATv2 equivariance proof); (iii) the FRAP action tokens $\{\mathbf{z}_{i,a}\}$ are node-equivariant by shared-weight computation; (iv) the actor applies the same MLP independently per valid phase; and (v) the factored critic aggregates from a local unordered neighbor set, which is permutation-invariant. The complete FGSv3 forward pass therefore satisfies node-permutation equivariance, enabling parameter sharing across all $N$ intersections.

\paragraph{Convergence disclaimer.}
The FGSv3 system inherits the convergence disclaimer of Section~\ref{sec:Appendix_convergence_disclaimer}. The soft Bellman contraction (Theorem~\ref{thm:bellman_contraction}) and soft policy improvement (Proposition) hold at the algorithm level and are unaffected by the architectural changes to the communication branch or the critic structure. The practical convergence of FGSv3 on ingolstadt21 is empirically assessed through training curves and validation metrics (Section~\ref{sec:ResultsAndDiscussion_Convergence}).

% ─────────────────────────────────────────────────────────────────────────────
\section{Soft Bellman Contraction and Policy Improvement}
\label{sec:Appendix_convergence}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Soft Bellman Contraction}
\label{sec:Appendix_bellman_contraction}

For a fixed policy $\pi$, the soft Bellman evaluation operator is
\begin{equation}
  (\mathcal{T}^\pi Q)(s,a)
  = r(s,a) + \gamma\,\mathbb{E}_{s'}\!\left[
    \sum_{a'}\pi(a'\mid s')\Bigl(Q(s',a') - \alpha\log\pi(a'\mid s')\Bigr)\right].
\end{equation}

\begin{theorem}[Soft Bellman contraction]
\label{thm:bellman_contraction}
$\mathcal{T}^\pi$ is a $\gamma$-contraction in $\|\cdot\|_\infty$:
\begin{equation}
  \|\mathcal{T}^\pi Q_1 - \mathcal{T}^\pi Q_2\|_\infty \leq \gamma\,\|Q_1-Q_2\|_\infty.
\end{equation}
Repeated application converges to the unique soft Q-function $Q^\pi_{\mathrm{soft}}$.
\end{theorem}
\begin{proof}
The entropy terms cancel between $Q_1$ and $Q_2$. For any $(s,a)$:
\begin{align*}
  \bigl|(\mathcal{T}^\pi Q_1 - \mathcal{T}^\pi Q_2)(s,a)\bigr|
  &= \gamma\left|\mathbb{E}_{s'}\!\left[\sum_{a'}\pi(a'\mid s')
     (Q_1(s',a')-Q_2(s',a'))\right]\right| \\
  &\leq \gamma\,\mathbb{E}_{s'}\!\left[\sum_{a'}\pi(a'\mid s')\|Q_1-Q_2\|_\infty\right]
   = \gamma\,\|Q_1-Q_2\|_\infty.
\end{align*}
Taking the supremum over $(s,a)$ and applying the Banach fixed-point theorem completes the proof.
\end{proof}

\subsection{Soft Policy Improvement}
\label{sec:Appendix_policy_improvement}

Given $Q^\pi$, define the updated policy $\pi_{\mathrm{new}}(a\mid s) = \exp(Q^\pi(s,a)/\alpha)\,/\,Z(s)$ where $Z(s)=\sum_b\exp(Q^\pi(s,b)/\alpha)$.

\begin{proposition}[Soft policy improvement]
$V^{\pi_{\mathrm{new}}}_{\mathrm{soft}}(s) \geq V^\pi_{\mathrm{soft}}(s)$ for all $s$.
\end{proposition}
\begin{proof}
Since $\log\pi_{\mathrm{new}}(a|s) = Q^\pi(s,a)/\alpha - \log Z(s)$, we have $Q^\pi(s,a)-\alpha\log\pi_{\mathrm{new}}(a|s)=\alpha\log Z(s)$ for all $a$, so $V^{\pi_{\mathrm{new}}}_{\mathrm{soft}}(s)=\alpha\log Z(s)$. Then:
\begin{align*}
  V^\pi_{\mathrm{soft}}(s)
  &= \sum_a\pi(a|s)\bigl[Q^\pi(s,a)-\alpha\log\pi(a|s)\bigr] \\
  &= \alpha\log Z(s)
     - \alpha\underbrace{D_{\mathrm{KL}}\!\bigl(\pi(\cdot|s)\,\|\,\pi_{\mathrm{new}}(\cdot|s)\bigr)}_{\geq\,0}
   \leq \alpha\log Z(s)
   = V^{\pi_{\mathrm{new}}}_{\mathrm{soft}}(s).
\end{align*}
\end{proof}

\subsection{Convergence Disclaimer}
\label{sec:Appendix_convergence_disclaimer}

The results above apply to the finite tabular entropy-regularized MDP. The implemented FGS/FGSv2 system departs from this setting through nonlinear function approximation, off-policy experience replay, simultaneous multi-agent learning, partial observability, target networks, and nonstationarity from co-adapting agent policies. These factors break the classical contraction argument, and global convergence to an optimal policy is \emph{not} guaranteed. The proofs establish the theoretical foundation of the SAC update targets; empirical convergence is assessed through training curves, TD errors, entropy, and validation metrics (Chapter~\ref{ch:ResultsAndDiscussion}).

% ─────────────────────────────────────────────────────────────────────────────
\section{Agent Communication Graph Extraction}
\label{sec:Appendix_TLSGraphExtraction}
% ─────────────────────────────────────────────────────────────────────────────

Algorithm~\ref{alg:tls_graph} describes how the agent communication graph $\mathcal{G}=(\mathcal{V},\mathcal{E},\mathbf{W})$ is derived from a SUMO \texttt{.net.xml} file. The core idea — contracting the full road network into a graph whose nodes are traffic-light controllers and whose edges are shortest-path connections between them, follows the super-edge construction. The extracted graphs for all six evaluation scenarios are shown in Figure~\ref{fig:scenario_topologies}.

\begin{algorithm}[ht]
  \caption{TLS Communication Graph Extraction from SUMO \texttt{.net.xml}}
  \label{alg:tls_graph}
  \begin{algorithmic}[1]
    \Require SUMO network file \texttt{net.xml}
    \Ensure Graph $\mathcal{G}=(\mathcal{V},\mathcal{E},\mathbf{W})$ with $\mathcal{V}$ = TLS nodes, $\mathcal{E}$ = super-edges, $w_{ij}=1/\tau_{ij}$

    \Statex \textbf{Phase 1 — Parse network}
    \State Parse all junctions with \texttt{tlLogic} entries $\Rightarrow$ TLS set $\mathcal{V}$; record 2-D positions
    \State Parse all non-internal road edges $\Rightarrow$ edge catalog $(e\!: \text{source}, \text{target}, \ell_e, \tau_e, \text{lanes})$
    \State Parse \texttt{connection} elements $\Rightarrow$ legal turn transitions $e \to e'$

    \Statex \textbf{Phase 2 — Build super-edges}
    \For{each TLS node $u \in \mathcal{V}$}
    \For{each outgoing road edge $e_0$ from $u$}
    \State Run Dijkstra on the legal-transition graph, starting at $e_0$, cost $= \sum \tau_e$
    \State \textbf{stop} at the first edge whose \emph{target} node $v \in \mathcal{V}$, $v \ne u$
    \If{$v$ found}
    \State Reconstruct path $\pi=(e_0,\ldots,e_k)$; compute $\ell = \sum_e \ell_e$, $\tau = \sum_e \tau_e$
    \State Record super-edge $(u,v,\tau,\pi)$; keep shortest-$\tau$ path if $(u,v)$ seen before
    \EndIf
    \EndFor
    \EndFor

    \Statex \textbf{Phase 3 — Assign edge weights}
    \For{each undirected super-edge $\{u,v\} \in \mathcal{E}$}
    \State $w_{uv} \mathrel{+}= 1/\tau_{uv}$ \quad (accumulate if multiple parallel paths exist)
    \EndFor
    \State Insert all edges bidirectionally to yield a symmetric graph

    \State \Return $\mathcal{G}=(\mathcal{V},\mathcal{E},\mathbf{W})$; serialize to JSON for training
  \end{algorithmic}
\end{algorithm}

The Dijkstra traversal operates on \emph{road edges} (not junctions) as nodes, following legal turn connections as transitions. This ensures only physically reachable paths are considered. Travel time $\tau_e = \ell_e / v_{\max,e}$ is computed from lane length and speed limit; when multiple parallel routes connect the same pair of TLS nodes their weights $1/\tau$ accumulate additively, reflecting the combined capacity of the corridor.

% ─────────────────────────────────────────────────────────────────────────────
\section{Statistical Notes on Single-Seed Results}
\label{sec:Appendix_stats}
% ─────────────────────────────────────────────────────────────────────────────

The RESCO benchmark specifies five random seeds with result averaging. Resource constraints limited most runs in this work to a single seed; results should be interpreted as indicative point estimates rather than statistically conclusive claims. For reference, a five-seed comparison would report mean $\bar x = S^{-1}\sum_s x_s$, standard error $\mathrm{SE}=\hat\sigma/\!\sqrt{S}$, and a 95\% confidence interval $\bar x \pm t_{S-1,\,0.975}\,\hat\sigma/\!\sqrt{S}$. Method comparisons benefit from paired differences $d_s = x_s^A - x_s^B$, which cancel seed-level variance: $\bar d \pm t_{S-1,\,0.975}\,\hat\sigma_d/\!\sqrt{S}$. Future work extending to five seeds will report paired-difference confidence intervals following this protocol.

\chapter{Full Experimental Results}
\label{ch:Appendix_FullResults}

This appendix presents complete numerical results across all six German RESCO scenarios and all evaluated methods, formatted after the style of the original RESCO supplementary material \citep{ault2021reinforcement}. Scenarios are classified into three scale tiers following the RESCO benchmark nomenclature:
\begin{itemize}
  \item \textbf{Single-intersection} (\textit{Single}): ingolstadt1 ($N=1$) and cologne1 ($N=1$).
  \item \textbf{Corridor} (\textit{Corr.}): ingolstadt7 ($N=7$, arterial) and cologne3 ($N=3$, arterial).
  \item \textbf{Regional} (\textit{Reg.}): ingolstadt21 ($N=21$) and cologne8 ($N=8$).
\end{itemize}

All results from this work are \textit{best-validation} values: the validation checkpoint with minimum mean waiting time observed during training, evaluated on a fixed held-out demand seed. Results marked $\circ$ indicate scenarios where the method was not yet run or crashed before producing valid metrics; these cells are candidates for re-run. Results marked $\dagger$ are from a partially trained run (FGS(MLP+GATv2+SAC) on Ing.~Reg.\ terminated after 17 of 200 training episodes due to resource exhaustion) and should not be compared directly to converged results. Results marked $\ddagger$ are early-training transients (best checkpoint at a very early episode with a single eval episode); their wait and queue metrics are internally inconsistent and are excluded from Bold comparisons. RESCO reference values (IDQN, IPPO, FMA2C) are reproduced from the original paper \citep{ault2021reinforcement} and represent best-episode averages over five random seeds.

\providecommand{\na}{$\circ$}        % missing / not run
\providecommand{\pt}{$^\dagger$}     % partial run
\providecommand{\pdd}{$^\ddagger$}   % early-training transient

% ─────────────────────────────────────────────────────────────────────────────
%  Table A.1  –  Baseline methods
% ─────────────────────────────────────────────────────────────────────────────
\begin{sidewaystable}[p]
\centering
\caption{Performance of baseline methods across all six RESCO scenarios (best-validation metrics, single seed unless noted). Columns follow the Single / Corridor / Regional tier classification. Reference results for IDQN, IPPO, and FMA2C are from \citet{ault2021reinforcement} (5-seed averages). FixedTime and MaxPressure are 5-episode static baselines from this work. DQN, PPO, IndSAC, CoLight, and FRAP are single-seed learning baselines from this work (best-validation checkpoint). Within each scenario column, \textbf{bold} marks the best value and \underline{underline} marks the second-best value per metric (lower is better for all metrics). $\circ$ = not run or not available.}
\label{tab:appendix_baselines}
\scriptsize
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{llrrrrrr}
\toprule
\textbf{Method} & \textbf{Metric}
  & \textbf{Ing.\ Single} & \textbf{Ing.\ Corr.}  & \textbf{Ing.\ Reg.}
  & \textbf{Col.\ Single} & \textbf{Col.\ Corr.}  & \textbf{Col.\ Reg.} \\
\midrule
\multicolumn{8}{l}{\textit{RESCO benchmark reference (5 seeds, best episode) \citep{ault2021reinforcement}}} \\[2pt]
\multirow{4}{*}{IDQN}
  & Avg.\ Delay     & \textbf{21.48} & \textbf{31.19} &  \textbf{61.11} &  26.05 &  24.55 &  22.06 \\
  & Avg.\ Trip Time & \textbf{35.29} & 68.69 & \textbf{198.72} &  43.59 &  59.00 &  86.02 \\
  & Avg.\ Wait      &  \textbf{3.93} &  \textbf{8.71} &  \textbf{21.36} &   7.98 &   8.50 &   5.46 \\
  & Avg.\ Queue     &  \textbf{0.43} &  0.67 &   \textbf{0.82} &   \textbf{2.09} &   0.87 &   0.38 \\[4pt]
\multirow{4}{*}{IPPO}
  & Avg.\ Delay     & \underline{21.85} & \underline{32.00} & 369.34 & 209.49 &  58.98 &  \underline{21.62} \\
  & Avg.\ Trip Time & 35.68 & 69.38 & 509.28 & 224.56 &  93.70 &  \underline{85.71} \\
  & Avg.\ Wait      &  \underline{4.20} &  \underline{8.86} & 315.31 & 193.66 &  43.50 &   5.19 \\
  & Avg.\ Queue     &  \underline{0.49} &  0.74 &   5.45 &  21.86 &   5.73 &   0.36 \\[4pt]
\multirow{4}{*}{FMA2C}
  & Avg.\ Delay     & 27.74 & 51.78 &  \underline{96.67} &  30.82 &  32.58 &  34.44 \\
  & Avg.\ Trip Time & 41.40 & 87.75 & \underline{233.04} &  48.07 &  66.83 &  98.68 \\
  & Avg.\ Wait      &  8.74 & 23.32 &  \underline{48.88} &  11.87 &  15.84 &  14.70 \\
  & Avg.\ Queue     &  1.07 &  1.86 &   1.84 &   3.20 &   2.01 &   1.03 \\
\midrule
\multicolumn{8}{l}{\textit{Static baselines --- this work (5-episode validation, no learning)}} \\[2pt]
\multirow{4}{*}{FixedTime}
  & Avg.\ Delay     &  40.8 & 122.9 & 145.7 &  54.6 &  63.7 &  51.5 \\
  & Avg.\ Trip Time &  54.4 & 135.5 & 281.8 &  65.6 &  93.9 & 115.7 \\
  & Avg.\ Wait      &  20.3 &  66.1 &  96.4 &  28.7 &  43.6 &  30.3 \\
  & Avg.\ Queue     &  7.36 &  5.16 &  3.17 & 15.53 &  5.68 &  2.18 \\[4pt]
\multirow{4}{*}{MaxPressure}
  & Avg.\ Delay     &  25.3 & 212.7 & 214.4 &  24.9 &  56.4 & 257.9 \\
  & Avg.\ Trip Time &  38.5 &  \textbf{66.9} & 235.5 &  42.5 &  81.6 & 231.4 \\
  & Avg.\ Wait      &   6.7 &  12.5 &  68.4 &   6.3 &  31.8 & 159.5 \\
  & Avg.\ Queue     &  1.30 &  \underline{0.49} &  1.23 &  3.12 &  3.37 & 10.58 \\
\midrule
\multicolumn{8}{l}{\textit{Learning baselines --- this work (200 episodes, best-validation checkpoint)}} \\[2pt]
\multirow{4}{*}{DQN}
  & Avg.\ Delay     &  22.1 &  34.4 & 115.1 &  26.0 &  22.2 &  25.6 \\
  & Avg.\ Trip Time &  36.0 &  71.7 & 253.3 &  43.6 &  58.6 &  89.8 \\
  & Avg.\ Wait      &   4.8 &   9.4 &  67.0 &   6.9 &   8.1 &   6.7 \\
  & Avg.\ Queue     &  1.15 &  0.81 &  1.12 &  3.49 &  1.32 &  0.46 \\[4pt]
\multirow{4}{*}{PPO}
  & Avg.\ Delay     &  21.9 &  42.3 & 211.1 &  \textbf{22.1} &  \underline{19.4} &  \textbf{20.8} \\
  & Avg.\ Trip Time &  \underline{35.6} &  77.1 & 236.2 &  \textbf{39.8} &  \underline{55.6} &  \textbf{85.0} \\
  & Avg.\ Wait      &   4.8 &  14.7 &  58.4 &   \textbf{4.8} &   \underline{6.0} &   \textbf{4.3} \\
  & Avg.\ Queue     &  0.93 &  0.95 &  \underline{0.96} &  \underline{2.43} &  \underline{0.83} &  \textbf{0.30} \\[4pt]
\multirow{4}{*}{IndSAC}
  & Avg.\ Delay     &  22.3 & 217.8 &  \na  &  \underline{23.0} &  \textbf{18.9} &  21.8 \\
  & Avg.\ Trip Time &  36.0 &  \underline{68.1} &  \na  &  \underline{40.7} &  \textbf{55.1} &  86.0 \\
  & Avg.\ Wait      &   4.9 &  13.1 &  \na  &   \underline{5.1} &   \textbf{5.5} &   \underline{4.7} \\
  & Avg.\ Queue     &  1.08 &  \textbf{0.44} &  \na  &  2.74 &  \textbf{0.77} &  \underline{0.33} \\[4pt]
\multirow{4}{*}{CoLight}
  & Avg.\ Delay     &  22.4 &  47.0 & 216.8 & \na   &  21.9 &  51.8 \\
  & Avg.\ Trip Time &  36.5 &  78.6 & 302.1 & \na   &  57.7 & 105.7 \\
  & Avg.\ Wait      &   5.3 &  17.5 & 120.8 & \na   &   7.9 &  23.8 \\
  & Avg.\ Queue     &  1.52 &  1.22 &   2.55 & \na   &  1.00 &  1.63 \\[4pt]
\multirow{4}{*}{FRAP}
  & Avg.\ Delay     &  26.6 &  \na  &  \na  & \na   &  39.8 &  44.9 \\
  & Avg.\ Trip Time &  40.1 &  \na  &  \na  & \na   &  71.1 & 108.6 \\
  & Avg.\ Wait      &   7.3 &  \na  &  \na  & \na   &  18.6 &  17.1 \\
  & Avg.\ Queue     &  1.67 &  \na  &  \na  & \na   &  2.78 &  1.20 \\
\bottomrule
\end{tabular}
\end{sidewaystable}

% ─────────────────────────────────────────────────────────────────────────────
%  Table A.2  –  FGS framework variants
% ─────────────────────────────────────────────────────────────────────────────
\begin{sidewaystable}[p]
\centering
\caption{Performance of FGS framework variants across all six RESCO scenarios (best-validation metrics, single seed). Encoder: MLP = two-layer MLP; FRAP = phase-competition encoder \citep{zheng2019learning}. Graph: GAT = dot-product attention; GATv2 = dynamic additive attention \citep{brody2022how}. RL: SAC = discrete soft actor-critic \citep{christodoulou2019soft}; PPO = proximal policy optimization \citep{schulman2017proximal}. $\circ$ = not yet run; $\dagger$ = terminated early (17\,/\,200 episodes); $\ddagger$ = early-training transient (single eval episode; wait and queue mutually inconsistent; see Table~\ref{tab:main_results} notes).}
\label{tab:appendix_fgs}
\scriptsize
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{llrrrrrr}
\toprule
\textbf{Variant} & \textbf{Metric}
  & \textbf{Ing.\ Single} & \textbf{Ing.\ Corr.}  & \textbf{Ing.\ Reg.}
  & \textbf{Col.\ Single} & \textbf{Col.\ Corr.}  & \textbf{Col.\ Reg.} \\
\midrule
\multicolumn{8}{l}{\textit{MLP encoder + SAC}} \\[2pt]
\multirow{4}{*}{FGS(MLP+GAT+SAC)}
  & Avg.\ Delay     &  22.6 &  \na  & \na   & \na   & \na   &  23.5 \\
  & Avg.\ Trip Time &  35.8 &  \na  & \na   & \na   & \na   &  87.8 \\
  & Avg.\ Wait      &   4.8 &  \na  & \na   & \na   & \na   &   5.5 \\
  & Avg.\ Queue     &  0.94 &  \na  & \na   & \na   & \na   &  0.38 \\[4pt]
\multirow{4}{*}{FGS(MLP+GATv2+SAC)}
  & Avg.\ Delay     &  23.0 &  \na  & 1201.8\pt & 23.0 & 20.8 & 22.2 \\
  & Avg.\ Trip Time &  36.8 &  \na  &  875.2\pt & 40.6 & 56.9 & 86.4 \\
  & Avg.\ Wait      &   5.5 &  \na  &  757.5\pt &  5.4 &  7.4 &  4.9 \\
  & Avg.\ Queue     &  1.08 &  \na  &  13.48\pt & 2.63 & 0.85 & 0.34 \\
\midrule
\multicolumn{8}{l}{\textit{MLP encoder + PPO}} \\[2pt]
\multirow{4}{*}{FGS(MLP+GAT+PPO)}
  & Avg.\ Delay     &  \na  &  \na  & 492.0 & \na   & \na   &  24.9 \\
  & Avg.\ Trip Time &  \na  &  \na  & 406.6 & \na   & \na   &  89.1 \\
  & Avg.\ Wait      &  \na  &  \na  & 261.2 & \na   & \na   &   6.7 \\
  & Avg.\ Queue     &  \na  &  \na  &  4.93 & \na   & \na   &  0.47 \\[4pt]
\multirow{4}{*}{FGS(MLP+GATv2+PPO)}
  & Avg.\ Delay     & 18.0\pdd & 26.1  & 234.9 & 12.5\pdd & 6.9\pdd  &  24.0 \\
  & Avg.\ Trip Time & 27.9\pdd & 66.5  & 253.5 & 26.4\pdd & 37.7\pdd &  88.3 \\
  & Avg.\ Wait      &  2.7\pdd &  6.8  &  88.6 &  1.1\pdd &  0.1\pdd &   6.3 \\
  & Avg.\ Queue     & 2.54\pdd &  0.50 &  2.58 & {---}\pdd & {---}\pdd &  0.44 \\
\midrule
\multicolumn{8}{l}{\textit{FRAP encoder + SAC}} \\[2pt]
\multirow{4}{*}{FGS(FRAP+GAT+SAC)}
  & Avg.\ Delay     & 590.7 &  \na  &  \na  & \na\rlap{$^*$} & 1270.4 & 222.9 \\
  & Avg.\ Trip Time &  95.0 &  \na  &  \na  & \na\rlap{$^*$} &  479.0 & 169.5 \\
  & Avg.\ Wait      &  78.1 &  \na  &  \na  & \na\rlap{$^*$} &  460.8 &  87.6 \\
  & Avg.\ Queue     &  2.55 &  \na  &  \na  & \na\rlap{$^*$} &  34.72 &  6.11 \\
\midrule
\multicolumn{8}{l}{\textit{FRAP encoder + PPO}} \\[2pt]
\multirow{4}{*}{FGS(FRAP+GATv2+PPO)}
  & Avg.\ Delay     &  \na  &  \na  & 617.0 & \na   & \na   &  24.6 \\
  & Avg.\ Trip Time &  \na  &  \na  & 614.8 & \na   & \na   &  88.9 \\
  & Avg.\ Wait      &  \na  &  \na  & 460.7 & \na   & \na   &   6.7 \\
  & Avg.\ Queue     &  \na  &  \na  &  7.77 & \na   & \na   &  0.47 \\
\midrule
\multicolumn{8}{l}{\textit{FGS generation 2 (action-token interface fix)}} \\[2pt]
\multirow{4}{*}{FGSv2-SAC}
  & Avg.\ Delay     & 17.4\pdd &  \na  & 53.7\pdd  & 12.0\pdd &  \na   &  23.6 \\
  & Avg.\ Trip Time & 27.4\pdd &  \na  & 191.4\pdd & 33.7\pdd &  \na   &  87.9 \\
  & Avg.\ Wait      &  2.1\pdd &  \na  &  13.9\pdd &  0.4\pdd &  \na   &   6.0 \\
  & Avg.\ Queue     &  2.55\pdd &  \na  &  12.90\pdd & {---}\pdd &  \na  &  0.40 \\
\midrule
\multicolumn{8}{l}{\textit{FGS generation 3 (demand-aware signal + factored critic)}} \\[2pt]
\multirow{4}{*}{FGSv3-PPO}
  & Avg.\ Delay     &  \na  &  \na  &  68.4 &  \na  &  \na  &  21.6 \\
  & Avg.\ Trip Time &  \na  &  \na  & 203.9 &  \na  &  \na  &  86.0 \\
  & Avg.\ Wait      &  \na  &  \na  &  22.9 &  \na  &  \na  &   4.8 \\
  & Avg.\ Queue     &  \na  &  \na  &  0.71 &  \na  &  \na  &  0.33 \\[4pt]
\multirow{4}{*}{FGSv3-SAC}
  & Avg.\ Delay     & 17.4\pdd &  \na  & 45.3\pdd  & 12.0\pdd &  \na  &  27.9 \\
  & Avg.\ Trip Time & 27.4\pdd &  \na  & 179.2\pdd & 33.7\pdd &  \na  &  92.1 \\
  & Avg.\ Wait      &  2.1\pdd &  \na  &  11.1\pdd &  0.4\pdd &  \na  &   6.0 \\
  & Avg.\ Queue     & 2.55\pdd &  \na  & 12.41\pdd & {---}\pdd &  \na  &  0.42 \\
\bottomrule
\end{tabular}
\vspace{6pt}

\footnotesize
$\circ$ Not yet run.\quad
$\dagger$ Partial: 17 of 200 training episodes completed (resource exhaustion).\quad
$\ddagger$ Early-training transient (single eval episode; wait and queue mutually inconsistent; excluded from comparison).\quad
$*$ Col.\ Single FGS(FRAP+GAT+SAC) run crashed before producing valid metrics.\quad
FGSv2-SAC and FGSv3-SAC produce identical metrics on Single-intersection scenarios ($N=1$): the GATv2 graph module has no neighbors to aggregate, so both generations reduce to the same single-agent controller.
\end{sidewaystable}

% ─────────────────────────────────────────────────────────────────────────────
%  Table A.3  –  Throughput summary
% ─────────────────────────────────────────────────────────────────────────────
\begin{sidewaystable}[p]
\centering
\caption{Vehicle throughput (arrived / departed, \%) across all methods and scenarios. Higher is better. $\circ$ = not run; $\dagger$ = partial run (17 episodes). RESCO reference algorithms did not report throughput; cells for IDQN, IPPO, FMA2C are therefore omitted.}
\label{tab:appendix_throughput}
\small
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lrrrrrr}
\toprule
\textbf{Method}
  & \textbf{Ing.\ Single} & \textbf{Ing.\ Corr.}  & \textbf{Ing.\ Reg.}
  & \textbf{Col.\ Single} & \textbf{Col.\ Corr.}  & \textbf{Col.\ Reg.} \\
\midrule
\multicolumn{7}{l}{\textit{Static baselines (this work; throughput metric not logged)}} \\[2pt]
FixedTime         & \na  & \na  & \na  & \na   & \na  & \na  \\
MaxPressure       & \na  & \na  & \na  & \na   & \na  & \na  \\
\multicolumn{7}{l}{\textit{Learning baselines (this work)}} \\[2pt]
DQN               & \na  & 97.6 & 94.1 & \na   & \na  & 98.5 \\
PPO               & 99.0 & 97.6 & \na  & \na   & 98.9 & 98.5 \\
IndSAC            & \na  & \na  & \na  & \na   & \na  & \na  \\
CoLight           & 99.1 & 97.6 & 93.2 & \na   & 98.8 & 97.9 \\
FRAP              & 99.1 & \na  & \na  & \na   & 97.7 & 97.8 \\
\midrule
\multicolumn{7}{l}{\textit{FGS variants (this work)}} \\[2pt]
FGS(MLP+GAT+SAC)    & 99.2 & \na  & \na       & \na   & \na   & 98.5 \\
FGS(MLP+GATv2+SAC)  & 99.0 & \na  & 43.0\pt   & 99.2  & 98.9  & 98.5 \\
FGS(MLP+GAT+PPO)    & \na  & \na  & 77.2      & \na   & \na   & 98.5 \\
FGS(MLP+GATv2+PPO)  & \na  & \na  & 90.9      & \na   & \na   & 98.4 \\
FGS(FRAP+GAT+SAC)   & 96.2 & \na  & \na       & \na\rlap{$^*$} & 62.1 & 95.2 \\
FGS(FRAP+GATv2+PPO) & \na  & \na  & 60.7      & \na   & \na   & 98.5 \\
\bottomrule
\end{tabular}
\vspace{4pt}

\footnotesize
$\circ$ Not yet run or throughput metric not available for this run.\quad $\dagger$ Partial run.\quad $*$ Crashed before producing valid metrics.
\end{sidewaystable}
