% Research Proposal Draft — LLNCS format
\documentclass[runningheads]{llncs}

\usepackage{graphicx}
\usepackage{url}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{microtype}
\usepackage{algorithm}
\usepackage{algpseudocode}

\begin{document}

\title{Flow-Coherent Communication Topologies for\\
Multi-Agent Reinforcement Learning-Based\\
Traffic Signal Control:\\
An InfoMap-Based Partitioning Framework}

\titlerunning{InfoMap-Based Communication for MARL-TSC}

\author{Duong Tran H.\inst{1}}

\authorrunning{D. Tran H.}

\institute{%
  Center for AI Research (CAIR), VinUniversity, Vietnam\\
  \email{duong.th@vinuni.edu.vn}%
}

\maketitle

\begin{abstract}
Traffic signal control~(TSC) in urban networks requires coordinated decisions
across hundreds of intersections, a challenge well-addressed by multi-agent
reinforcement learning~(MARL). Agent communication is central to coordination
quality, yet existing MARL-TSC methods determine communication topology from
geometric proximity or learned reward signals---neither of which reflects the
underlying vehicle flow structure. We propose using the map
equation~(InfoMap)~\cite{rosvall2008maps} to partition road networks by
origin-destination~(OD) demand coherence, yielding communication topologies
in which within-module agent pairs have higher mutual information about future
traffic states. We operationalise this idea at three distinct levels: regional
reward shaping~(IDQN/IPPO), regional critic design~(MAPPO), and regional
communication graph~(CoLight-style regional GATs). A rigorous evaluation
protocol following the \emph{rliable}~\cite{agarwal2021rliable} methodology
specifies sample-efficiency metrics~(area under the learning curve, episodes to
threshold) with stratified bootstrap confidence intervals and performance
profiles. We further propose an attention range analysis---grounded in a
principled range measure for graph neural network operators~\cite{lrd2025}---to
mechanistically validate that InfoMap partitioning focuses attention within
flow-coherent modules. A synthesis of three converging bodies of
literature---MARL-TSC communication protocols, network partitioning for traffic
management, and flow-based community detection---reveals a clear gap: no
existing study applies flow-based community detection to define MARL
communication neighbourhoods in TSC. The information bottleneck~(IB) framework
widely applied in MARL communication~\cite{ding2021imac} provides motivating
grounding: InfoMap provides a domain-motivated communication prior that reduces
the topology search space that IB methods must otherwise explore from task
reward signals alone---though the formal equivalence between the two objectives
remains a conjecture requiring proof. We identify five research gaps, formalise
a testable hypothesis, and specify a complete multi-stage experimental design
for empirical validation.
\keywords{%
  multi-agent reinforcement learning \and
  traffic signal control \and
  communication topology \and
  InfoMap \and
  map equation \and
  network partitioning \and
  origin-destination flow \and
  information bottleneck \and
  sample efficiency%
}
\end{abstract}

% ============================================================
\section{Introduction}
% ============================================================

Urban traffic congestion imposes substantial economic and environmental costs.
Traffic signal control---the real-time management of intersection signal
phases---is among the most tractable levers available to transportation
engineers, yet optimal coordination across an urban network remains an open
problem. With $n$ signalised intersections, the joint action space grows
exponentially, making centralized optimal control computationally infeasible at
city scale.

Multi-agent reinforcement learning reframes the problem by assigning one agent
per intersection, each learning a local policy. Decentralisation makes the
approach scalable but introduces partial observability---each agent observes
only local traffic state---and the need for coordination. Communication between
agents is the primary mechanism for resolving partial observability, and the
structure of that communication (who communicates with whom, what is
communicated, and when) directly determines coordination quality.

The dominant approach to communication structure in MARL-TSC is geographic:
agents communicate with their $k$-hop road-network neighbours. This is
intuitively reasonable but conflates two distinct design objects.
The \emph{road network topology}---the physical graph of intersections and
roads---determines what communication is \emph{possible}. The
\emph{communication topology}---which agent pairs actually exchange
information---determines what coordination \emph{occurs}. Existing methods
set these equal by default, but vehicle flow does not respect road-network
adjacency. A vehicle entering intersection $A$ may traverse several
intermediate intersections before reaching its true destination zone, making
immediate road-neighbours of $A$ poor predictors of its downstream impact.
The right communication topology should reflect \emph{where vehicles go},
not \emph{which roads connect}.

We formalise the TSC problem as a Networked Distributed POMDP
(ND-POMDP)~\cite{nair2005ndpomdp,oliehoek2012ndpomdp}, in which the
\emph{coordination graph}---the graph over agents whose edges determine which
pairs share non-separable joint utility---is the central design object.
ND-POMDP theory establishes that computational tractability scales inversely
with coordination graph density~\cite{kok2006ndpomdp}: a sparser graph that
correctly identifies the non-separable pairs achieves both better coordination
and lower computational cost than a dense graph that includes spurious edges.

We investigate whether the map equation
(InfoMap)~\cite{rosvall2008maps} provides a principled criterion for
designing this coordination graph from origin-destination~(OD) flow data.
InfoMap identifies communities where a random walker tends to stay; in traffic,
the walker is a vehicle and communities correspond to OD demand basins. The
core hypothesis is that agents within the same InfoMap module have higher
mutual information about each other's future traffic states, and therefore
constitute the non-separable pairs in the coordination graph.  We further
conjecture that InfoMap modules correspond to \emph{spill-back propagation
  zones}: congestion that spills back from one intersection tends to propagate
within the module rather than across flow-basin boundaries~\cite{dependencydyn2025}.

\medskip\noindent\textbf{Contributions.}
\begin{enumerate}
  \item A formal ND-POMDP reframing of the MARL-TSC communication problem,
  separating the \emph{road network topology} (physical infrastructure)
  from the \emph{coordination graph} (communication design object),
  and establishing why OD flow is the principled criterion for
  coordination graph design.
  \item An information-theoretic argument connecting the map equation to
        MARL communication topology, formalised as two testable hypotheses:
        H1~(within-module mutual information advantage) and H2~(spill-back
        containment within modules).
        \item A comprehensive literature synthesis across four bodies of work
        (MARL-TSC, communication topology design, network partitioning,
        and spatio-temporal dependency), identifying six concrete research gaps.
  \item A three-level experimental protocol operationalising the InfoMap
        partition at the reward level~(IDQN/IPPO), critic level~(MAPPO),
        and communication graph level~(regional GATs), with a factorial
        ablation design isolating each level's contribution and a temporal
        sensitivity ablation addressing OD demand stationarity.
  \item A rigorous evaluation methodology based on the rliable
        framework~\cite{agarwal2021rliable}---interquartile mean~(IQM),
        performance profiles, and stratified bootstrap confidence
        intervals---addressing reproducibility concerns in MARL-TSC evaluation.
  \item An attention range analysis protocol---grounded in the range measure for
        graph neural network operators~\cite{lrd2025}---to mechanistically
        validate that regional GAT partitioning sharpens attention focus
        within flow-coherent modules.
\end{enumerate}

% ============================================================
\section{Background}
\label{sec:background}
% ============================================================

\subsection{Decentralised Control and the Coordination Graph}

Traffic signal control with $n$ intersections is naturally modelled as a
\textbf{Decentralised Partially Observable Markov Decision Process}
(Dec-POMDP)~\cite{oliehoek2012ndpomdp}: each agent $i$ observes only its
local state $o_i$ (queue lengths, phase durations), takes a local action $a_i$
(signal phase), and receives a local reward $r_i$. The joint reward $R = \sum_i
  r_i$ is the global objective (minimise average travel time).

For large $n$, the Dec-POMDP is intractable in general (NEXP-hard).
The \textbf{Networked Distributed POMDP (ND-POMDP)}~\cite{nair2005ndpomdp,kok2006ndpomdp} introduces structure by restricting which agents interact:
the reward factorises as
\begin{equation}
  R(s,a) = \sum_{(i,j)\in E_C} \rho_{ij}(s_i, s_j, a_i, a_j)
  \label{eq:ndpomdp}
\end{equation}
where $\mathcal{G}_C = (V, E_C)$ is the \emph{coordination graph} and
$\rho_{ij}$ is a local payoff function between agent pair $(i,j)$.
Crucially, \textbf{tractability scales inversely with graph density}:
sparser $\mathcal{G}_C$ supports exact or near-exact solutions that become
infeasible on the full graph~\cite{kok2006ndpomdp}.

This makes the \emph{design of $\mathcal{G}_C$} the central algorithmic
question in scalable MARL-TSC. Existing work implicitly sets
$E_C = E_{\text{road}}$ (the road network edges), i.e.\ agents coordinate
iff they are road-neighbours. We argue this conflates two distinct objects:
\begin{itemize}
  \item \textbf{Road network topology} $G_{\text{road}} = (V, E_{\text{road}})$:
        fixed physical infrastructure; determines what communication is
        \emph{possible}.
  \item \textbf{Coordination graph} $\mathcal{G}_C = (V, E_C)$:
        the design variable; determines what coordination \emph{occurs}.
\end{itemize}
Setting $E_C = E_{\text{road}}$ is the default but not the optimum.
The optimal $\mathcal{G}_C$ connects exactly those agent pairs $(i,j)$ for
which $\rho_{ij}$ is non-negligible---i.e.\ pairs whose traffic states
are predictably coupled.  In a TSC network, these are the pairs through
which vehicles actually travel together: OD-coherent pairs.

\subsection{The Map Equation and InfoMap}

The map equation~\cite{rosvall2008maps,rosvall2009mapequation} provides the
theoretical foundation. The core insight is that community structure in a
network can be identified by finding the partition that best compresses the
description of a random walk using the minimum description length~(MDL)
principle:

\begin{equation}
  L(\mathcal{M}) = q_{\curvearrowright} H(\mathcal{Q})
    + \sum_{i=1}^{k} p_i^{\circlearrowleft} H(\mathcal{P}_i)
  \label{eq:mapequation}
\end{equation}

\noindent where $q_{\curvearrowright}$ is the rate at which a random walker
moves \emph{between} modules, $H(\mathcal{Q})$ is the entropy of
between-module movement, $p_i^{\circlearrowleft}$ is the rate of staying
\emph{within} module $i$, and $H(\mathcal{P}_i)$ is the entropy of
within-module movement. Minimising $L(\mathcal{M})$ over all partitions
$\mathcal{M}$ identifies communities where the walker---and in traffic,
vehicles---tend to remain.

\medskip\noindent\textbf{Key algorithmic properties}~\cite{edler2023community}:
\begin{itemize}
  \item Handles directed, weighted edges without discarding direction (critical
        for traffic OD matrices);
  \item Does not require specifying the number of communities $k$ in advance;
  \item Does not suffer from the resolution limit affecting modularity-based
        methods (Louvain, Girvan-Newman);
  \item Produces a hierarchical codebook---a natural two-level communication
        vocabulary.
\end{itemize}

Applied to a traffic network, nodes are intersections and edge weights
represent directed OD flow between them. The partition that minimises
Eq.~\eqref{eq:mapequation} recovers intersections that vehicles tend to visit
together---OD-coherent demand basins.

\medskip\noindent\textbf{Scope of the random walk approximation.}
The map equation assumes a memoryless random walk; real vehicle routing is
destination-directed and influenced by GPS and habitual paths. Applied to
aggregated intersection-pair flow counts from simulation or historical data,
the partition approximates OD demand basins rather than strict random walk
communities. This is a purposeful approximation: we use flow volume as a proxy
for flow coherence, and whether this is sufficient for defining effective MARL
communication topologies is precisely what Hypothesis~H1
(Section~\ref{sec:theory}) is designed to test empirically.

% ============================================================
\section{Related Work}
\label{sec:related}
% ============================================================

\subsection{MARL-TSC Communication Protocols}

\textbf{CoLight}~\cite{wei2019colight} applies graph attention networks to
MARL-TSC, selectively weighting messages from road-graph neighbours.
Its key limitation is that the communication graph equals the road graph,
ignoring vehicle flow structure. We address this by partitioning the global
GAT into regional GATs aligned with InfoMap modules.

\textbf{Representational communication}~\cite{bokade2023representational} shows
that message quality limits large-scale TSC coordination; \textbf{CVDMARL}~\cite{cvdmarl2024}
confirms structured communication outperforms independent learning.
\textbf{HiLight}~\cite{zhu2026hilight} and \textbf{Li et al.}~\cite{li2025topology}
show that region definition and explicit topological structure beyond the road
graph improve MARL performance, though neither uses OD flow as the partitioning signal.
\textbf{RegionLight}~\cite{regionlight2024} and \textbf{NeurComm}~\cite{neurcomm}
further confirm neighbourhood scope shapes coordination quality.

\textbf{STMARL}~\cite{stmarl2020} decomposes coordination into spatial
(road-adjacency GNN) and temporal (RNN history) dimensions, finding that
\emph{spatial structure boosts TSC performance more than temporal dependency}.
This calibrates our priority: InfoMap addresses the spatial coordination
dimension, the higher-leverage target.

\textbf{Dependency Dynamics in TSC}~\cite{dependencydyn2025} shows that
inter-agent coordination is necessary \emph{if and only if} spill-back
congestion connects two agents' queues---when no spill-back occurs, independent
agents achieve near-optimal reward. This provides a traffic-theoretic grounding
for $E_C$: edges should connect exactly those pairs between which spill-back
can propagate. We conjecture InfoMap modules correspond to spill-back zones
(Hypothesis~H2, Section~\ref{sec:theory}).

\subsection{Communication Topology Design in MARL}

\textbf{ATOC}~\cite{jiang2018atoc} shows that selective communication via
attentional gating outperforms full broadcast.
\textbf{I2C}~\cite{i2c2021} learns per-pair communication value but lacks
scalability and domain-structure grounding.
\textbf{Correlated topology learning}~\cite{aamas2021correlated} optimises
the communication graph as a latent variable, yielding dynamic but opaque
and non-transferable topologies.
\textbf{ExpoComm}~\cite{expocomm2025} demonstrates that pre-designed structural
topologies ($O(\log n)$ diameter exponential graphs) match or exceed learned
topologies at scale, directly motivating domain-prior approaches.
Three TSC-specific properties justify a domain prior: (i)~stable urban demand
makes an OD-based prior informative from episode one; (ii)~early-policy topologies
can drift from converged-policy optima; (iii)~InfoMap partitions transfer to
new networks via OD data alone, without MARL retraining.

\textbf{IB methods} formalise the communication design question.
IMAC~\cite{ding2021imac} achieves ${>}80\%$ message reduction while retaining
task-relevant MI; IB-VQ~\cite{sun2026ibvq} extends this to bandwidth-efficient
discrete coding; Heterogeneous IB Coordination Graphs~\cite{hetib2025}
co-design graph structure and per-edge bandwidth.
InfoMap provides a domain-motivated topology prior that reduces the search space
IB methods must explore, though formal equivalence between the map equation
objective and the IB objective remains a conjecture.

\textbf{GNN dependency analysis}~\cite{lrd2025} quantifies how much a graph
model's output at node $v$ changes when input features of node $u$---$h$ hops
away---are perturbed, via aggregated absolute Jacobians, yielding per-hop
influence profiles $\bar{T}_h$.
We adapt this framework to MARL-TSC (Section~\ref{sec:depquant}) as a
post-hoc diagnostic for whether trained critic and communication models
concentrate spatial dependency within InfoMap modules.

\subsection{Network Partitioning for Traffic Signal Control}

\textbf{Laplacian partitioning}~\cite{laplacian2009ieee} identifies signal
coordination groups spectrally, but from road-network topology rather than flow.
\textbf{OD-flow partitioning}~\cite{odflow2019partition} demonstrates that
flow-derived partitions outperform adjacency-based methods on delay and
speed---the strongest empirical precedent for our hypothesis.
\textbf{MFD-based partitioning}~\cite{mfdpartition2020,communitydetection2021plosone}
targets density homogeneity for perimeter control, not OD flow coherence.

\textbf{Feudal MARL with Adaptive Partition}~\cite{ma2022feudal} is the closest
prior work, using GNN+MCTS to partition the network for a feudal hierarchy.
Key differences: criterion is reward-driven (not flow-based OD); partition
defines a policy hierarchy (not a communication topology); information-theoretic
partition properties are not analysed.

\textbf{GPLight}~\cite{liu2023gplight} groups agents by observation MI for
parameter sharing rather than communication topology design; clusters are
learned during training, not pre-computed from domain data.

\subsection{InfoMap in Urban Mobility and Temporal Stability}

\textbf{Lucchini et al.}~\cite{lucchini2023intraod} apply InfoMap to hourly
intra-urban OD networks and find that 82.9\% of regions maintain stable
community membership across time windows; only peripheral boundary nodes
shift between modules. This supports a time-aggregated OD partition as a
valid coordination graph approximation.

\textbf{Flow stability theory}~\cite{flowstability2022} extends community
detection to temporal networks, identifying dynamical scales at which community
boundaries stabilise. In urban traffic, communities are stable within a
regime (peak/off-peak) but may shift across regimes, grounding the temporal
sensitivity ablation in Section~\ref{sec:temporal}.

\medskip\noindent\textbf{Critical gap.} No existing study applies the InfoMap
codebook to define MARL communication neighbourhoods in TSC, nor examines
how OD temporal variation affects static partition quality across RL training episodes.

\subsection{Rigorous Evaluation in Deep Reinforcement Learning}

Henderson et al.~\cite{henderson2018matters} show that deep RL performance
differences often disappear under different seeds or protocols---a concern
amplified in MARL-TSC, where most results rely on single-seed final-episode comparisons.
Agarwal et al.~\cite{agarwal2021rliable} formalise \emph{rliable}: interquartile
mean (IQM) for efficient point estimates, stratified bootstrap confidence intervals,
performance profiles, and probability of improvement.
Our protocol adopts all four measures to ensure sample-efficiency claims about
InfoMap partitioning are statistically defensible.

% ============================================================
\section{Theoretical Framework}
\label{sec:theory}
% ============================================================

\subsection{The Coordination Graph as the Central Design Object}
\label{sec:coordgraph}

Building on the ND-POMDP formulation~(Eq.~\eqref{eq:ndpomdp}), the
central design question is: \emph{which pairs $(i,j)$ should have an edge
  in $\mathcal{G}_C$?}  We decompose the factors influencing this choice:

\medskip\noindent\textbf{Topology vs.\ partition.}
\begin{itemize}
  \item \textbf{Global topology} (road network): specifies the
        \emph{physical reachability} of agents---what communication
        infrastructure exists. This is fixed and given.
  \item \textbf{Partition} (coordination grouping): specifies the
        \emph{functional communication structure}---which pairs should
        communicate because their joint utility $\rho_{ij}$ is non-trivial.
        This is the design variable.
\end{itemize}
Both affect distinct aspects of what agents learn.  The partition
affects \emph{who} receives the reward, credit, or message signal
(the ``who communicates'' question).  Within each level of the
three-level protocol, it shapes what the RL agent \emph{sees and acts upon}:
at Level~1 the reward signal, at Level~2 the value function input, at
Level~3 the attention neighbourhood.

\medskip\noindent\textbf{Map equation as coordination description length.}
The map equation $L(\mathcal{M})$ (Eq.~\eqref{eq:mapequation}) quantifies
the average number of bits needed to describe a random walk on the
OD flow graph under partition $\mathcal{M}$.  Two terms capture exactly
the coordination cost structure of ND-POMDP:
\begin{itemize}
  \item $q_{\curvearrowright} H(\mathcal{Q})$: the code cost of
        \emph{between-module} transitions---analogous to the cost of
        cross-module coordination edges in $\mathcal{G}_C$.
  \item $\sum_i p_i^{\circlearrowleft} H(\mathcal{P}_i)$: the code cost
        of \emph{within-module} transitions---analogous to the cost of
        intra-module coordination.
\end{itemize}
The InfoMap optimum $\mathcal{M}^* = \arg\min L(\mathcal{M})$ minimises
total description length by \emph{maximising within-module flow retention
  at minimum cross-module code cost}.  Translated to coordination graph design:
$\mathcal{M}^*$ identifies the partition where retaining within-module
coordination edges captures the most flow-relevant agent dependencies at
minimum graph density cost.  This is the first formal connection, to our
knowledge, between the map equation and ND-POMDP coordination graph design.

\subsection{Spatio-Temporal Dependency Decomposition}
\label{sec:spatiotemporal}

Traffic signal control involves two orthogonal dependency structures that
both affect coordination quality~\cite{stmarl2020}:

\medskip\noindent\textbf{Spatial dependency.}
At a given time $t$, the state $X_i(t)$ of intersection $i$ is influenced
by the states of physically connected intersections via vehicle flow.
This is the dependency targeted by the coordination graph $\mathcal{G}_C$:
if $(i,j) \in E_C$, agents $i$ and $j$ share signals that carry spatially
correlated traffic information.  The InfoMap partition determines the
\emph{spatial scope} of coordination.

\medskip\noindent\textbf{Temporal dependency.}
Each agent's future state $X_i(t+\tau)$ depends on its own history
$\{X_i(t'), t' < t\}$ (signal phase memory, queue accumulation).
This is the dependency targeted by recurrent architectures and observation
history windows in the RL policy.  The partition does not directly control
this dimension; it is handled by the agent's internal state representation.

\medskip\noindent\textbf{OD temporal variability.}
A third form of temporal dependency concerns the \emph{OD demand pattern
  itself}: the flow matrix $W_{ij}$ changes across time of day (morning peak,
off-peak, evening peak).  This determines whether a \emph{static} InfoMap
partition---estimated from one OD snapshot---remains valid across all training
episodes.  STMARL~\cite{stmarl2020} finds that spatial dependency boosts
TSC performance more than temporal dependency in most networks, suggesting that
even an approximate static partition may capture the dominant coordination
structure.  Lucchini et al.~\cite{lucchini2023intraod} provide empirical
support: 82.9\% of urban regions maintain stable community membership across
hours, with only peripheral boundary nodes shifting between regimes.
Flow stability theory~\cite{flowstability2022} refines this: communities are
stable within a traffic \emph{regime} (peak or off-peak), but may differ
across regimes.  The temporal sensitivity ablation (Section~\ref{sec:temporal})
directly tests this boundary.

\medskip\noindent\textbf{Why the 1-hour OD snapshot is a defensible starting
  point.}  Within any 1-hour simulation window at a stable demand regime, OD
flow is approximately stationary in the Wardrop sense: holding demand constant,
vehicles settle into a quasi-equilibrium assignment.  The partition computed
from this window captures the dominant spatial coordination structure for that
regime.  The question of whether it transfers to other regimes is not whether
the partition is ``wrong'' in absolute terms, but whether regime-specific
partitions improve over the single static partition---a difference addressable
by the C3/C4 temporal ablation conditions in Section~\ref{sec:temporal}.

\subsection{Testable Hypotheses}
\label{sec:hypotheses}

Let $X_i(t)$ denote the traffic state at node $i$ at time $t$ and
$\mathcal{M} = \{M_1, \ldots, M_k\}$ be an InfoMap partition.
The framework rests on a main hypothesis supported by two complementary
sub-hypotheses at different levels of abstraction: H2 is the
traffic-theoretic \emph{mechanism}, H1 is its information-theoretic
\emph{consequence}. Together they imply three operational predictions
at the coordination level (Section~\ref{sec:depquant}).

\medskip\noindent\textbf{Main Hypothesis~H (Partition Quality).}
An OD-flow-coherent partition (InfoMap) achieves higher dependency
retention---$\Psi(\mathcal{P}_{\mathrm{InfoMap}}, \mathbf{D}) >
  \Psi(\mathcal{P}_{\mathrm{METIS}}, \mathbf{D})$---for the same
number of modules $K$, leading to better coordination at all three
levels of the protocol.

\medskip\noindent\textbf{Hypothesis H1 (Spatial MI Advantage).}
For any intersections $i, j$ in the same module $M_k$ and $l \notin M_k$,
within-module pairs share higher mutual information about each other's
\emph{spatial} traffic state---the cross-agent dimension indexed by $(i,j)$,
not the within-agent temporal dimension:
\begin{equation}
  I\!\left(X_i(t);\, X_j(t+\tau)\right) >
  I\!\left(X_i(t);\, X_l(t+\tau)\right),
  \quad \text{on average over propagation lags } \tau \in [1, T].
  \label{eq:h1}
\end{equation}
H1 is \emph{motivated by}, but not formally derived from, the map equation:
if the OD flow graph approximates a random walk, vehicles tend to remain
within $M_k$, so congestion at $i$ propagates preferentially to $j \in M_k$
rather than to $l \notin M_k$.
\textbf{H1 is treated as an empirical hypothesis, not a derived theorem.}

\medskip\noindent\textbf{Hypothesis H2 (Spill-back Containment).}
Spill-back congestion is more likely to stay within an InfoMap module
than to cross module boundaries:
\begin{equation}
  \Pr\!\left(\text{spill-back}_{i \to j} \mid i,j \in M_k\right)
  > \Pr\!\left(\text{spill-back}_{i \to l} \mid i \in M_k,\, l \notin M_k\right).
  \label{eq:h2}
\end{equation}
H2 is grounded in dependency dynamics~\cite{dependencydyn2025}: coordination is
necessary \emph{if and only if} spill-back connects two agents' queues.
If InfoMap modules correspond to spill-back zones, $E_C$ restricted to
within-module edges captures exactly the necessary coordination pairs.
\textbf{H2 is the causal mechanism for H1}: if spill-back stays within
modules, within-module agents' states co-vary via shared congestion
propagation, raising $I(X_i; X_j)$ above the cross-module baseline.
Unlike H1, H2 is directly testable from queue and trajectory data.

\medskip\noindent\textbf{Verification protocol.}
Before MARL training, collect $\sim$1000 warm-up steps with a random policy.
\begin{enumerate}
  \item \emph{H1}: Compute $\hat{I}(X_i(t); X_j(t+\tau))$ via KSG estimator
  at $\tau \in \{1, 5, 15, 30\}$ steps for all pairs $(i,j)$.
  Classify as within-module or cross-module.
  Test with a Wilcoxon signed-rank test ($p < 0.05$); visualise as
  a $|V|\times|V|$ MI heatmap with InfoMap boundaries overlaid.
  \item \emph{H2}: Extract spill-back events from vehicle trajectory data
        (queue lengths exceeding junction capacity at consecutive intersections).
        Test the fraction of within-module spill-back vs.\ the
        base rate with a one-sample proportion test.
        \end{enumerate}
        Failure modes and implications: if both H1 and H2 fail, halt training and
        revise the partition criterion.
        If H1 fails but H2 holds, spill-back containment still partially justifies
        the partition.
        If H1 holds but H2 fails, the MI advantage is present but the spill-back
        mechanism is incorrect---revise the mechanism, not the partition.
        
        \noindent Section~\ref{sec:depquant} operationalises H1 and H2 through a
        dependency quantification framework, introducing the metrics needed to
        test~H and deriving per-level operational predictions from H1 and H2.
        
        \subsection{Spatial Dependency Metrics and Three-Level Predictions}
        \label{sec:depquant}
        
        H1 and H2 predict that within-module pairs exhibit stronger spatial coupling
        than cross-module pairs. To connect this prediction to each coordination level,
        we need a common dependency metric that (i)~is computable without presupposing
        the partition, (ii)~yields a scalar per pair $(i,j)$, and (iii)~can be evaluated
        under any candidate partition via a single partition-quality score.
        
        \medskip\noindent\textbf{Two complementary instruments.}
        
        \noindent\textit{Physical dependency (model-free):}
        H1 directly provides a dependency score---the empirical mutual information
        between agents' traffic states:
        \begin{equation}
          D^{\mathrm{MI}}_{ij} = \hat{I}\!\left(X_i(t);\, X_j(t{+}\tau)\right)
          \label{eq:dmi}
        \end{equation}
        estimated by the KSG estimator on warm-up trajectories.
        $D^{\mathrm{MI}}_{ij}$ is model-free and applies at Level~1: regional reward
        averaging (Eq.~\eqref{eq:regionalreward}) transmits a correlated gradient
        signal within modules when within-module $D^{\mathrm{MI}}_{ij}$ is high,
        reducing variance relative to independent reward.
        Note that at Level~1 (IPPO/IDQN), agents act independently---there is no
        cross-agent input path in the policy network, so a Jacobian-based measure
        is not applicable.
        
        \noindent\textit{Learned dependency (Jacobian on the global model):}
        For Levels~2 and~3, we adapt the Jacobian-based influence framework
        of~\cite{lrd2025} to the MARL setting.
        For a model $F$ trained \emph{without} partition constraints (the global model),
        define the spatial influence of agent~$j$ on agent~$i$ as:
        \begin{equation}
          D_{i \leftarrow j}(F)
          = \sum_{a,k}
          \left|\frac{\partial F_{i,a}}{\partial X_{j,k}}\right|
          \label{eq:djac}
        \end{equation}
        where $X_{j,k}$ is feature~$k$ of agent~$j$'s observation and $F_{i,a}$ is
        output dimension~$a$ of~$F$ at agent~$i$.
        The \emph{global-model} restriction is essential: a partitioned model has no
        cross-module input pathway by construction, making its cross-module Jacobian
        zero by definition---a structural artefact, not evidence of low dependency.
        We instantiate Eq.~\eqref{eq:djac} at two levels:
        \begin{align*}
          D^{Q}_{i \leftarrow j}
           & = D_{i \leftarrow j}(V^{\mathrm{global}})
           &                                                          & \text{(global MAPPO critic, Level~2)}          \\
          D^{\mathrm{G}}_{i \leftarrow j}
           & = D_{i \leftarrow j}(\mathbf{h}_i^{(L),\mathrm{global}})
           &                                                          & \text{(global CoLight GAT embedding, Level~3)}
        \end{align*}
        
        \medskip\noindent\textbf{Hop-distance profile.}
        For either instrument, aggregate by road-graph hop distance $\rho(i,j)$:
        \begin{equation}
          \bar{D}_h = \frac{1}{|V|}
          \sum_{i}\sum_{j:\,\rho(i,j)=h} D_{i \leftarrow j}
          \label{eq:barDh}
        \end{equation}
        This is the MARL analog of the per-hop influence $\bar{T}_h$ of~\cite{lrd2025}:
        $\bar{D}_h$ quantifies the average spatial dependency from agents exactly
        $h$ road-graph hops away.
        A steeply decaying $\bar{D}_h$ confirms that local (1--2 hop) coordination
        suffices; a slowly decaying profile reveals multi-hop coupling---and specifies
        the minimum module radius for effective coordination.
        
        \medskip\noindent\textbf{Dependency retention ratio.}
        Given a fixed dependency matrix $\mathbf{D}$ and a partition
        $\mathcal{P}=\{P_1,\ldots,P_K\}$, define:
        \begin{align}
          \text{IntraDep}(\mathcal{P},\mathbf{D})
           & = \sum_{k}\sum_{i,j \in P_k} D_{ij},
          \label{eq:intradep}                                                     \\
          \Psi(\mathcal{P},\mathbf{D})
           & = \frac{\text{IntraDep}(\mathcal{P},\mathbf{D})}{\sum_{i,j} D_{ij}}.
          \label{eq:psi}
        \end{align}
        $\Psi \in [0,1]$ is the \emph{dependency retention ratio}: the fraction of
        total spatial dependency captured within partition boundaries.
        Computing $\Psi$ on the same $\mathbf{D}$ under InfoMap vs.\ METIS tests
        whether the flow-based partition aligns better with the dependency structure,
        without circularity.
        
        \medskip\noindent\textbf{Why InfoMap should maximise $\Psi$.}
        The map equation maximises the fraction of vehicle-flow steps retained within
        modules. High within-module vehicle flow implies~(H1) that queue states co-vary
        via shared arrivals, and~(H2) that spill-back congestion stays within module
        boundaries---both raising $D_{ij}$ for within-module pairs.
        Accordingly, InfoMap should achieve higher $\Psi$ than METIS on both the MI
        matrix and the global-model Jacobian matrices, for the same $K$.
        
        \medskip\noindent\textbf{Operational predictions from H1 and H2.}
        These follow directly from the causal chain H2$\to$H1$\to$H and are the
        primary empirical targets of the experimental design.
        \begin{enumerate}
          \item \textbf{OP1 (L1, reward):} H1 implies $D^{\mathrm{MI}}_{ij}$ is
                systematically higher within InfoMap modules, hence
                $\Psi(\mathcal{P}_{\mathrm{InfoMap}}, D^{\mathrm{MI}}) >
                  \Psi(\mathcal{P}_{\mathrm{METIS}}, D^{\mathrm{MI}})$.
                Regional reward averaging then operates on correlated signals,
                reducing gradient noise and training instability.
                \emph{Directional support}: IPPO$+$flow shows 3.2$\times$ fewer
                instability spikes than IPPO$+$METIS on Cologne8 (Table~\ref{tab:prelim}).
          \item \textbf{OP2 (L2, critic):} H1 implies $D^{Q}_{i \leftarrow j}$
                concentrates within InfoMap modules, hence
                $\Psi(\mathcal{P}_{\mathrm{InfoMap}}, D^Q) >
                  \Psi(\mathcal{P}_{\mathrm{METIS}}, D^Q)$.
                The regional critic $V_k(\{o_i\}_{i \in M_k})$ captures dominant
                value dependencies without modelling irrelevant cross-module states,
                reducing credit-assignment variance.
                \emph{Directional support}: MAPPO$+$flow achieves 2.85--3.99$\times$
                lower ATT than MAPPO$+$METIS on Ingolstadt21.
          \item \textbf{OP3 (L3, communication):} H1 implies $D^{\mathrm{G}}_{i \leftarrow j}$
                concentrates within InfoMap modules, hence
                $\Psi(\mathcal{P}_{\mathrm{InfoMap}}, D^{\mathrm{G}}) >
                  \Psi(\mathcal{P}_{\mathrm{METIS}}, D^{\mathrm{G}})$,
                observed as a lower influence-weighted range $\hat{\rho}_u$
                (Eq.~\eqref{eq:rangemeasure}, Section~\ref{sec:attnrange})
                for regional vs.\ global CoLight, provided module size
                is sufficient (${\geq}3$--$4$ agents per module).
        \end{enumerate}

\subsection{The Codebook as Communication Vocabulary}

InfoMap produces a hierarchical two-level codebook~\cite{rosvall2011multilevel}
that maps naturally to a two-level communication protocol:

\begin{enumerate}
  \item \textbf{Module codewords} (Level 2): compact regional identity
        (which demand basin an agent belongs to).
  \item \textbf{Within-module codewords} (Level 1): agent position within
        its region.
\end{enumerate}

This induces a two-level message structure:
agents within the same module exchange full state vectors
(within-module codewords), while agents in different modules exchange only
module-level aggregate representations (module codewords)---reducing
cross-module communication bandwidth while preserving intra-module
coordination fidelity.

\subsection{Comparison of Topology Design Approaches}

Table~\ref{tab:topology} compares existing topology designs across three
properties relevant to our hypothesis.

\begin{table}[t]
  \centering
  \caption{Comparison of communication topology design criteria.}
  \label{tab:topology}
  \small
  \begin{tabularx}{\linewidth}{l X c c c}
    \toprule
    \textbf{Method} & \textbf{Partition Criterion} & \textbf{Interpretable}
      & \textbf{Flow-Grounded} & \textbf{Transferable} \\
    \midrule
    $k$-hop neighbours   & Road adjacency          & Yes     & No       & Yes \\
    ATOC~\cite{jiang2018atoc}
                         & Observation correlation & No      & No       & No  \\
    ExpoComm~\cite{expocomm2025}
                         & Graph diameter          & Yes     & No       & Yes \\
    GPLight~\cite{liu2023gplight}
                         & Observation MI          & Partial & No       & Partial \\
    Ma \& Wu~\cite{ma2022feudal}
                         & Reward signal           & No      & Indirect & No \\
    \textbf{InfoMap (proposed)}
                         & OD flow dynamics        & \textbf{Yes}
                                                            & \textbf{Yes}
                                                                       & \textbf{Yes} \\
    \bottomrule
  \end{tabularx}
\end{table}

The proposed approach is the only one that is simultaneously interpretable,
directly grounded in traffic flow, and transferable across networks without
MARL \emph{policy} retraining. \textit{Note on transferability:} ``transferable''
means the InfoMap partition can be recomputed from the new network's OD data
without any MARL training; the MARL policy itself still requires training on
the target network. Transfer of the \emph{partition structure} (zero-shot)
vs.\ transfer of the \emph{policy} (full retraining) are tested separately in
Section~\ref{sec:experiment}.

% ============================================================
\section{Research Gaps}
\label{sec:gaps}
% ============================================================

Synthesis of four literature bodies (MARL-TSC communication, coordination
graph theory, network partitioning, and spatio-temporal dependency) reveals
six research gaps (Table~\ref{tab:gaps}).
Gaps G1 and G2 are blocking for algorithm development; G3 and G6 are
theoretical; G4 and G5 address validity and methodological standards.

\begin{table}[t]
  \centering
  \caption{Research gaps identified from four-body literature synthesis.}
  \label{tab:gaps}
  \small
  \begin{tabularx}{\linewidth}{@{} l X p{1.25cm} X @{}}
    \toprule
    \textbf{Gap} & \textbf{Description} & \textbf{Sig.} & \textbf{Resolution} \\
    \midrule
    G1 & InfoMap partition not used to define MARL communication topology in TSC
        & High & Proposed framework (Section~\ref{sec:framework}) \\[3pt]
    G2 & No comparison of flow-based vs.\ geometric partition across all three coordination levels
        & High & Factorial study (Section~\ref{sec:experiment}) \\[3pt]
    G3 & No formal connection between map equation and ND-POMDP coordination graph; road topology conflated with coordination graph
        & High & ND-POMDP + map-equation formalisation (Section~\ref{sec:theory}) \\[3pt]
    G4 & Temporal sensitivity of static InfoMap partitions under regime shifts not studied in MARL-TSC
        & Medium & Temporal ablation, T0--T5 (Section~\ref{sec:temporal}) \\[3pt]
    G5 & MARL-TSC evaluations rarely apply multi-seed statistical protocols~\cite{henderson2018matters}
        & Medium & rliable with IQM + bootstrap CIs~\cite{agarwal2021rliable} \\[3pt]
    G6 & Coordination \emph{topology} (who communicates) and \emph{content} (what is communicated) never jointly studied under one partition criterion
        & Medium & Three-level protocol (Section~\ref{sec:framework}) \\
    \bottomrule
  \end{tabularx}
\end{table}

% ============================================================
\section{Proposed Framework}
\label{sec:framework}
% ============================================================

\subsection{Problem Formulation: ND-POMDP with InfoMap Coordination Graph}
\label{sec:formulation}

We model MARL-TSC as an ND-POMDP~\cite{nair2005ndpomdp,kok2006ndpomdp}
with the following components:
\begin{itemize}
  \item \textbf{Agents:} $\mathcal{N} = \{1, \ldots, n\}$, one per
        signalised intersection.
  \item \textbf{State:} $s = (s_1, \ldots, s_n)$ where $s_i$ includes
        queue lengths and current phase at intersection $i$.
  \item \textbf{Observations:} $o_i \sim O(s_i, a_i)$; each agent observes
        only its local state (partial observability).
  \item \textbf{Actions:} $a_i \in \mathcal{A}_i$, signal phase selection.
  \item \textbf{Factored reward:} $R(s,a) = \sum_i r_i(s_i, a_i)$ where
        $r_i = -\sum_l w_l(t)$ (negative sum of waiting times at intersection $i$).
        This is additively separable, so $\rho_{ij} \neq 0$ only when
        spill-back creates direct coupling between $i$ and $j$.
  \item \textbf{Coordination graph:} $\mathcal{G}_C = (V, E_C)$, designed
        via InfoMap partition of the OD flow graph.
\end{itemize}

The partition $\mathcal{M} = \{M_1, \ldots, M_k\}$ induced by InfoMap
on the OD flow graph defines the coordination graph:
\begin{equation}
  E_C = \{(i,j) : \exists m \text{ s.t.\ } i,j \in M_m\}
  \label{eq:ec}
\end{equation}
i.e.\ within-module pairs are connected; cross-module pairs are disconnected
(or connected only via a coarse inter-module aggregate channel).
This is the minimal ND-POMDP coordination graph consistent with the
OD demand structure, and by the tractability result of~\cite{kok2006ndpomdp}
it achieves the best computational-coordination trade-off among all
OD-partition-induced graphs.

\subsection{OD Flow Graph Construction}
\label{sec:odconstruction}

The InfoMap algorithm requires a directed weighted graph $G=(V,E,W)$ where
$W_{ij}$ is the vehicle flow volume from intersection $i$ to intersection $j$.
This is not a standard simulator output and requires explicit instrumentation.

\medskip\noindent\textbf{In CityFlow:} Enable vehicle trajectory logging via
the \texttt{roadnet.json} configuration. For each vehicle trip, extract the
ordered sequence of intersections traversed. For each consecutive pair
$(i, j)$ in the sequence, increment $W_{ij}$. After the warm-up period,
normalise by total vehicle count to obtain a directed flow-proportion matrix.

\medskip\noindent\textbf{In SUMO:} Use the \texttt{edgedata} output with
\texttt{tripinfo} to reconstruct intersection-pair traversal counts.

\medskip\noindent\textbf{Cold-start problem.} The warm-up phase (first
$\sim$1000 steps) uses a random or rule-based policy before MARL training
begins. The OD flow under a random policy may differ from the OD flow under
the converged MARL policy, because signal timing affects which routes vehicles
take. Two mitigations:
\begin{enumerate}
  \item \textbf{Historical OD data} (preferred when available): use real-world
        OD matrices from the target city rather than simulation warm-up, decoupling
        partition quality from warm-up policy.
  \item \textbf{Iterative re-partitioning}: after $N$ training epochs, collect
        fresh OD data from the partially-trained policy and re-run InfoMap.
        Repeat once (re-partitioning more than twice risks training instability).
\end{enumerate}
This limitation is acknowledged; its severity is assessed via the ablation
comparing warm-up-OD partition vs.\ historical-OD partition.

\subsection{InfoMap Partitioning and H1 Verification}

Before MARL training begins, the road network is partitioned and H1 is
verified on warm-up data (Algorithm~\ref{alg:infomap}).

\begin{algorithm}[H]
  \caption{InfoMap Communication Topology Construction}
  \label{alg:infomap}
  \begin{algorithmic}[1]
    \Require Directed weighted graph $G=(V,E,W)$ (from Section~\ref{sec:odconstruction});
             $V$=intersections, $W_{ij}$=OD flow volume
    \Ensure Partition $\mathcal{M}=\{M_1,\ldots,M_k\}$; codebook $C$;
            inter-module adjacency matrix $A^{\text{cross}}$
    \State \textbf{(Cold-start)} If historical OD available, use as $W$;
           else collect $W$ via random-policy warm-up (flag policy mismatch)
    \State Run \textsc{Infomap}$(G)$ $\to$ $\mathcal{M}$, codebook $C$
    \State \textbf{Assert non-degeneracy:} $1 < k < |V|$ and all
           $|M_i| \geq 2$ (degenerate partition $\Rightarrow$ check OD data)
    \State Verify H1: compute $\hat{I}(X_i;X_j)$ for within- vs.\ cross-module
           pairs via KSG estimator; test with Wilcoxon signed-rank
    \State \textbf{if} H1 fails \textbf{then} halt and revise hypothesis
    \State Construct $A^{\text{cross}}$: inter-module edges with aggregate flow
    \State \Return $\mathcal{M}$, $C$, $A^{\text{cross}}$
  \end{algorithmic}
\end{algorithm}

\subsection{Three-Level Application Protocol}
\label{sec:threelevels}

The InfoMap partition can be applied at three orthogonal levels within a
MARL-TSC system, each targeting a different inductive bias:

\medskip\noindent\textbf{Level 1 --- Regional reward shaping (IDQN / IPPO):}
Replace the single-intersection reward $r_i(t)$ with an intra-module average:
\begin{equation}
  \tilde{r}_i(t) = \alpha\, r_i(t) + (1-\alpha)\,
  \frac{1}{|M_k|} \sum_{j \in M_k} r_j(t), \quad i \in M_k.
  \label{eq:regionalreward}
\end{equation}
This encourages agents within the same flow basin to internalise each other's
congestion signals without requiring any explicit communication channel.
Pilot experiments (Section~\ref{sec:prelim}) reveal a \emph{cold-start
  instability} at Level~1: when $\alpha < 1$ from episode~1, the group
averaging corrupts the individual reward gradient during the early learning
phase, causing catastrophically high travel times for 400--1000 episodes.
We therefore add an \emph{annealing schedule}:
\begin{equation}
  \alpha(e) = 1 - (1-\alpha_{\min})\cdot\min\!\left(\frac{e}{E_{\text{warm}}}, 1\right)
  \label{eq:alphaanneal}
\end{equation}
starting from pure individual reward ($\alpha=1$) and linearly decreasing to
$\alpha_{\min}$ over the first $E_{\text{warm}}$ episodes.  This lets agents
first learn a stable individual policy before the regional coupling is
introduced.  The annealing ablation (C1-$\alpha$ condition in
Table~\ref{tab:factorial}) tests whether this mitigates the cold-start penalty.

\medskip\noindent\textbf{Level 2 --- Regional critic design (MAPPO):}
Replace the global centralised critic $V(s)$ with $k$ regional critics,
each receiving only the joint state of agents within module $M_k$:
$V_k\!\left(\{o_i\}_{i \in M_k}\right)$.
This reduces the critic's input dimensionality from $O(n)$ to $O(n/k)$,
improving credit assignment within flow-coherent groups while preserving
the CTDE training paradigm.

\medskip\noindent\textbf{Level 3 --- Regional communication graph (CoLight):}
Replace the single global Graph Attention Network with $k$ regional GATs,
one per InfoMap module. Each regional GAT computes attention only among
agents within its module. Cross-module coordination is handled by a lightweight
inter-module GAT operating on module-level aggregate representations.
This implements the two-level codebook structure from Section~\ref{sec:theory}
as a direct architectural change to the CoLight attention mechanism.

\medskip\noindent\textbf{Composability.} The three levels are orthogonal:
Level~1 affects reward signals, Level~2 affects value estimation, and Level~3
affects message passing. They can be applied independently or in combination.
The factorial ablation design in Section~\ref{sec:factorial} isolates the
contribution of each.

% ============================================================
\section{Experimental Design}
\label{sec:experiment}
% ============================================================

\subsection{Simulation Environments}

\noindent\textbf{Platform:} CityFlow or SUMO with LibSignal benchmark
wrappers~\cite{libsignal}, enabling reproducible comparison against published
baselines.

\noindent\textbf{Networks:}
\begin{itemize}
  \item Synthetic $6{\times}6$ grid (core experiment, all baselines);
  \item Synthetic $10{\times}10$ grid (scalability ablation);
  \item Real-world: Cologne8, Ingolstadt21 (RESCO standard benchmarks);
  \item Jinan $3{\times}4$, Manhattan segment (transferability test).
\end{itemize}

\noindent\textbf{Demand scenarios:}
uniform demand; AM/PM peak-hour OD patterns; mixed peak/off-peak.
Multiple demand scenarios on the same network produce different InfoMap
partitions (different $L(\mathcal{M})$ and intra-partition flow ratios
$\phi_k$), enabling regression analysis of which partition quality metrics
predict RL improvement (RQ3).

\noindent\textbf{Episode length:} 3600-second simulation with 15-second
phase intervals (CTDE framework: MAPPO or QMIX base algorithm).

\noindent\textbf{Seeds:} minimum 5 seeds per condition (defensible minimum);
8 seeds recommended; all seeds fixed and reported for reproducibility.

\subsection{Baseline Ladder and Sparsification Control}

Table~\ref{tab:baselines} defines the experimental conditions.
Baseline L1.5~(random sparse graph, same per-agent degree as the InfoMap
partition) is the critical control for the sparsification confound: if the
InfoMap condition (L4) beats L1.5, the benefit is attributable to
\emph{flow-coherent} topology selection, not merely to communication graph
sparsification. Without this control, any advantage of InfoMap over
$k$-hop neighbours is confounded between flow structure and reduced degree.

\begin{table}[t]
  \centering
  \caption{Experimental baseline ladder. L1.5 isolates the sparsification confound.}
  \label{tab:baselines}
  \small
  \begin{tabularx}{\linewidth}{l l X}
    \toprule
    \textbf{Level} & \textbf{Method} & \textbf{Partition Criterion} \\
    \midrule
    L0   & No communication (independent)  & --- \\
    L1   & $k=1$ hop neighbours             & Road adjacency \\
    L1.5 & Random sparse graph (same degree as InfoMap) & None (sparsification control) \\
    L2   & METIS partition                  & Graph balance (structural) \\
    L3   & GPLight MI clustering~\cite{liu2023gplight} & Observation MI \\
    L4   & \textbf{InfoMap partition (proposed)} & OD flow dynamics \\
    L5   & InfoMap + IB-VQ compression~\cite{sun2026ibvq} & OD flow + bandwidth opt.\\
    \bottomrule
  \end{tabularx}
\end{table}

\subsection{Factorial Ablation Design}
\label{sec:factorial}

To isolate which application level drives performance, we run a $2 \times 5$
factorial ablation: $\{$InfoMap, METIS$\}$ partition $\times$ five activation
patterns (Table~\ref{tab:factorial}).
\begin{table}[t]
  \centering
  \caption{Factorial ablation conditions. Each row applies the partition at
           a different combination of the three levels.}
  \label{tab:factorial}
  \small
  \begin{tabularx}{\linewidth}{l c c c X}
    \toprule
    \textbf{Condition} & \textbf{Reward (L1)}
      & \textbf{Critic (L2)} & \textbf{Comm (L3)} & \textbf{Purpose} \\
    \midrule
    C0 & off & off & off & Baseline (no partition) \\
    C1 & on  & off & off & Level~1 reward only \\
    C2 & off & on  & off & Level~2 critic only \\
    C3 & off & off & on  & Level~3 comm only \\
    C4 & on  & on  & on  & All levels combined \\
    \bottomrule
  \end{tabularx}
\end{table}
Conditions C1--C3 identify which level provides the greatest marginal benefit;
C4 tests whether the levels are complementary or redundant.
Run across $\{$MAPPO, CoLight$\}$ $\times$ $\{$Cologne8, Ingolstadt21$\}$
$\times$ 5 seeds = 100 runs total.
Pairwise comparisons use Wilcoxon signed-rank tests with Bonferroni correction
for the family of 10 pairwise contrasts.

\subsection{Rigorous Evaluation Methodology}
\label{sec:evalmethod}

Standard MARL-TSC evaluation reports final-episode ATT averaged across a small
number of runs, which is insufficient for statistically credible comparisons
\cite{henderson2018matters}. We adopt the rliable framework
\cite{agarwal2021rliable} throughout.

\medskip\noindent\textbf{Primary task metrics:} average travel time~(ATT,
seconds/vehicle; primary), average waiting time~(AWT), average queue
length~(AQL). ATT is the canonical comparison metric across RESCO and
LibSignal.

\medskip\noindent\textbf{Sample efficiency metrics} (the central claim):
\begin{itemize}
  \item \textbf{Area Under Learning Curve (AUC-LC):} $\int_0^N \text{ATT}(e)\,de$
        over $N$ training episodes via trapezoid rule. Lower AUC-LC means
        less cumulative delay incurred during training---directly measuring
        how quickly the agent learns to control traffic well.
  \item \textbf{Episodes to Threshold (ET-90, ET-95):} number of training
        episodes to reach 90\% and 95\% of the best observed ATT for that
        condition. ET-90 is more robust to final-episode noise; ET-95 provides
        a stricter benchmark.
  \item \textbf{Convergence plateau:} episode range where ATT standard
        deviation falls below 0.5\% of the mean, measuring training stability.
\end{itemize}

\medskip\noindent\textbf{Aggregate statistics (rliable protocol):}
All aggregate metrics are reported as IQM $\pm$ 95\% stratified bootstrap CI
(2000 bootstrap samples, stratified across seeds and environments). Pairwise
\emph{probability of improvement} $\Pr(\text{InfoMap} > \text{METIS})$ is
reported with 95\% CI. Performance profiles plot the fraction of (algorithm,
environment, seed) triples achieving at least $x$\% of the best observed ATT,
for $x \in [0,1]$, providing a full distributional view.

\medskip\noindent\textbf{Statistical tests:} Wilcoxon signed-rank test on
per-run AUC-LC scores for pairwise comparisons. Bonferroni correction applied
across the family of $k$ comparisons in the factorial design. Effect size
reported as rank-biserial correlation $r$.

\medskip\noindent\textbf{Partition quality regression (RQ3):} for each
network/demand scenario, compute intra-partition flow ratio $\phi_k$,
map equation value $L(\mathcal{M})$, graph conductance $\Phi$, and
NMI between InfoMap and METIS partitions. Regress $\Delta$AUC-LC
(InfoMap $-$ METIS) on these metrics to identify which partition property
predicts RL improvement. A significant positive coefficient on $\phi_k$
would support the claim that flow coherence, not merely graph structure,
drives the benefit.

\subsection{Temporal Partition Sensitivity Ablation}
\label{sec:temporal}

The OD flow matrix $W$ used to compute the InfoMap partition depends on
the demand pattern during the estimation window.  If episodes span multiple
demand regimes (e.g.\ a SUMO scenario with time-varying
\texttt{<flow>} rates), a single static partition may be suboptimal for
some episodes.  Flow stability theory~\cite{flowstability2022} and the
Lucchini et al.\ findings~\cite{lucchini2023intraod} bound the severity:
core communities ($\approx$83\% of nodes) are stable across regimes; only
boundary nodes shift.  We test this empirically with five conditions:

\begin{table}[t]
  \centering
  \caption{Temporal partition sensitivity ablation.  Each condition varies
    the OD source used to estimate the InfoMap partition while holding
    the MARL algorithm (MAPPO) and network (Ingolstadt21) constant.
    ``Cross-regime'' episodes are episodes where the active demand
    pattern differs from the partition's estimation regime.}
  \label{tab:temporal}
  \small
  \begin{tabularx}{\linewidth}{l X X}
    \toprule
    \textbf{Condition} & \textbf{OD source (partition)}          & \textbf{Tests}         \\
    \midrule
    T0                 & No partition (MAPPO baseline)           & Reference              \\
    T1                 & METIS (graph balance, no flow)          & Spatial without flow   \\
    T2                 & InfoMap from 1-hour off-peak OD         & Single-regime static   \\
    T3                 & InfoMap from 1-hour peak OD             & Regime specificity     \\
    T4                 & InfoMap from aggregated 24-hour OD      & Regime-averaged static \\
    T5                 & Dynamic re-partition every $N$ episodes & Temporal adaptation    \\
    \bottomrule
  \end{tabularx}
\end{table}

\noindent The critical comparison is T2 vs.\ T3 (does the partition regime
matter?) and T4 vs.\ T2/T3 (does aggregation degrade or improve?).  If
T2 $\approx$ T3 $\approx$ T4, the static 1-hour partition is sufficient and
the temporal limitation is not a material concern.  If T3 $\gg$ T2 on
peak-hour episodes, regime-specific partitioning is warranted.

\subsection{Attention Range Analysis}
\label{sec:attnrange}

This analysis implements the Level~3 instantiation of the spatial dependency
framework (Section~\ref{sec:depquant}).
We compute $D^{\mathrm{G}}_{i \leftarrow j}$ (Eq.~\eqref{eq:djac}) on the
\emph{global} CoLight GAT---without any partition constraints---to avoid
the trivially-zero cross-module Jacobians that arise in a partitioned model.
The influence-weighted range $\hat{\rho}_u$ defined below then summarises the
hop-distance distribution of that dependency.

\medskip\noindent\textbf{Range measure for GNN operators.}
Following~\cite{lrd2025}, the node-level \emph{range measure} derived from
the normalised Jacobian of the GNN output with respect to each input node is:
\begin{equation}
  \hat{\rho}_u(\mathbf{F}) = \mathbb{E}_{v \sim I_u}\!\left[d_G(u,v)\right]
  \label{eq:rangemeasure}
\end{equation}
where $I_u$ is the influence distribution of target node $u$ (computed as
the normalised Jacobian $\partial F_u / \partial \mathbf{x}_v$ over all
source nodes $v$), and $d_G(u,v)$ is the graph distance.
Intuitively, $\hat{\rho}_u$ is the expected distance from which node~$u$
receives task-relevant information through the GNN.
For the trained CoLight GAT (global and regional), we compute $\hat{\rho}$
as follows:
\begin{enumerate}
  \item For each target intersection $i$, compute the normalised Jacobian
  $s_{ij} = \partial F_i / \partial \mathbf{x}_j$
  from all other intersections $j$ in the trained model.
  \item Compute $\hat{\rho}_i = \sum_j s_{ij}\, d_G(i,j)$ (expected
  influence distance).  Average over all $i$ to get network-level
  $\hat{\rho}$.
  \item Compare $\hat{\rho}$ for global CoLight (no partition) vs.\ regional
  CoLight (InfoMap partition).  A lower $\hat{\rho}$ after partitioning
  indicates that attention is concentrating within modules rather than
  diffusing globally.
  \item Decompose $\hat{\rho}$ into within-module and cross-module components
  to verify that regional partitioning suppresses cross-module influence
  while preserving within-module attention.
\end{enumerate}

\medskip\noindent\textbf{Attention rollout visualisation.}
We additionally apply attention rollout~\cite{abnar2020rollout} to trace
cumulative influence through stacked attention layers backward from each
agent's decision, revealing which intersections' states actually drove
the action. Visualising rollout heatmaps before and after partitioning
shows whether attention ``leaks'' across module boundaries in the global GAT
but concentrates within modules in the regional GAT.

\medskip\noindent\textbf{Predicted outcome and connection to H1.}
If H1 holds---within-module agents carry higher mutual information about
each other's future states---we expect that a well-trained GAT will
\emph{discover} this structure by allocating higher attention weights to
within-module neighbours, even without explicit partition constraints.
The global CoLight should then exhibit a moderate $\hat{\rho}$ biased
toward module boundaries; the regional CoLight should exhibit a lower
$\hat{\rho}$ with sharper concentration.  Crucially, if the global CoLight
also shows module-aligned attention with $\hat{\rho}$ comparable to the
regional variant, this would suggest the partition is redundant and the
GAT self-organises to respect flow boundaries---a scientifically interesting
null result.  If the regional CoLight achieves both lower $\hat{\rho}$
\emph{and} better ATT, the partition provides structural inductive bias
that training alone cannot efficiently discover.

% ============================================================
\section{Preliminary Experimental Results}
\label{sec:prelim}
% ============================================================

We report results from a pilot study on two SUMO scenarios---Cologne8
(8~intersections, $k{=}3$) and Ingolstadt21 (21~intersections, $k{\in}\{4,5\}$)---to
provide a \emph{directional} signal ahead of the full experimental programme.
All runs used a single seed (seed~0); no statistical significance claims are made.
We track \texttt{eval/metrics/waitingTime} (average waiting time per vehicle; ATT),
lower is better.

\subsection{Cologne8 Results (1500 Episodes)}

Table~\ref{tab:prelim} summarises the six key runs for which full learning
curves were obtained (IPPO and MAPPO families); IDQN data is incomplete due
to a baseline crash.

\begin{table}[t]
  \centering
  \caption{Cologne8 pilot results (seed=0, $k{=}3$, 1500~episodes).
           AUC-LC~$=\int_0^{1500}\!\mathrm{ATT}(e)\,de$ (trapezoidal,
           $\times10^3$ (s/veh)$\cdot$ep; lower is better).
           ET$(\le t)$~= first episode at which ATT~$\le t$ s/veh.
           Late-$\sigma$~= std of ATT over episodes 1201--1500.}
  \label{tab:prelim}
  \small
  \begin{tabularx}{\linewidth}{l r r r r r r}
    \toprule
    \textbf{Condition} & \textbf{Final ATT} & \textbf{AUC-LC\,($k$)} &
    \textbf{Late-$\sigma$} & \textbf{ET(50)} & \textbf{ET(40)} & \textbf{ET(28)} \\
    \midrule
    IPPO baseline    & 39.2 & \phantom{0}66.7 & 1.2  & 240 & 690  & ---  \\
    IPPO + flow      & 27.1 & 237.6           & 1.0  & 490 & 550  & 1330 \\
    IPPO + METIS     & 33.8 & 509.2           & 68.5\rlap{$^*$} & 1350 & 1360 & --- \\
    \midrule
    MAPPO baseline   & 28.9 & 165.2           & 0.5  & 320 & 350  & 570  \\
    MAPPO + flow     & \textbf{26.0} & \textbf{94.4} & 0.5 & 270 & 290 & 540 \\
    MAPPO + METIS    & 26.8 & 109.3           & 0.5  & 270 & 300  & 560  \\
    \bottomrule
  \end{tabularx}
  {\footnotesize $^*$IPPO$+$METIS has not converged at ep~1500 (Late-$\sigma = 68.5$
   indicates ongoing oscillation of $\pm$135\,s/veh).}
\end{table}

\medskip\noindent\textbf{MAPPO (Level~2---Regional Critic).}
Both partition conditions improve upon the baseline in all metrics.
MAPPO$+$flow achieves a \textbf{10\%} reduction in final ATT (26.0 vs 28.9\,s/veh)
and a \textbf{43\% reduction in AUC-LC} (94.4k vs 165.2k), demonstrating
substantially better sample efficiency.  Notably, MAPPO$+$flow reaches the
baseline's best ATT level at episode~470---\textbf{31\% of the training budget}.
All three MAPPO variants converge to a zero-instability plateau (Late-$\sigma \le 0.5$),
demonstrating that Level~2 provides clean, stable training dynamics regardless of
partition choice.  MAPPO$+$flow leads MAPPO$+$METIS on AUC-LC by 13.7\% (94.4 vs 109.3\,k)
and final ATT by 0.8\,s/veh (26.0 vs 26.8), providing a directional signal consistent
with~H1.

\medskip\noindent\textbf{IPPO (Level~1---Regional Reward).}
The regional reward introduces a \emph{cold-start problem}: both partition variants
begin training at catastrophically high ATT ($\sim$1\,200--1\,300\,s/veh), compared to
the IPPO baseline's modest 128\,s/veh at episode~10.  This explains the dramatically
worse AUC-LC for IPPO$+$flow (237.6k) and IPPO$+$METIS (509.2k) relative to the
baseline (66.7k), despite eventually achieving \emph{better} final ATT.
After the cold-start transient, IPPO$+$flow reaches a stable plateau of 27.1\,s/veh
(31\% below baseline), but exhibits 26~instability spikes above 100\,s/veh after
episode~400 (worst: ATT$=$241\,s/veh at ep~650).  IPPO$+$METIS is far more
problematic, with 83~post-400 spikes (worst: 436\,s/veh), and has not
converged at episode~1500 (Late-$\sigma = 68.5$).

These patterns suggest two important design insights.  First, Level~1 (regional reward)
is the most sensitive to partition quality: the InfoMap partition's OD-coherence reduces
training instability relative to METIS (26 vs 83 spikes), consistent with H1---OD-coherent
modules share correlated future traffic states and thus produce a more consistent group
reward signal, making the averaged reward less noisy as a training signal.
The METIS partition, whose edges are determined by graph balance rather than flow
coherence, groups agents with less correlated states, increasing gradient noise.
Second, the cold-start instability (Section~\ref{sec:threelevels},
Eq.~\eqref{eq:alphaanneal}) is a real phenomenon: when the regional average is applied
from episode~1, the individual gradient is corrupted before the agent has a stable
baseline policy, causing the catastrophic early ATT.  The annealing schedule is now
added to the experimental plan as the C1-$\alpha$ ablation condition.

\medskip\noindent\textbf{Flow vs.\ METIS (Levels~1--2).}
Across all Cologne8 conditions where convergence was achieved, InfoMap (flow) consistently
outperforms METIS: final ATT advantage of 0.8\,s/veh for MAPPO and 6.7\,s/veh for IPPO;
AUC-LC advantage of 13.7k for MAPPO and 271.6k for IPPO; instability 3.2$\times$ lower
for IPPO.  This directionally supports H1 at Levels~1--2 on a single seed.

\medskip\noindent\textbf{Level~3 --- CoLight (Regional GAT).}
Table~\ref{tab:colight} shows the three CoLight variants on Cologne8, now fully
completed.  The result is unexpected: the flow partition \emph{diverges}, while METIS
is essentially neutral.

\begin{table}[htb]
  \centering
  \caption{Cologne8 CoLight results (seed=0, $k{=}3$, 1500~episodes).
           Late-mean and Late-$\sigma$ computed over episodes 1001--1500.}
  \label{tab:colight}
  \small
  \begin{tabularx}{\linewidth}{l r r r}
    \toprule
    \textbf{Condition} & \textbf{Eval ATT} & \textbf{Late-mean} & \textbf{Late-$\sigma$} \\
    \midrule
    CoLight (no partition) & \textbf{16.20} & 14.4 & 1.3 \\
    CoLight + METIS $k{=}3$ & 15.92 & 18.4 & 1.9 \\
    CoLight + flow $k{=}3$  & 28.06 & 56.2 & 25.3\rlap{$^*$} \\
    \bottomrule
  \end{tabularx}
  {\footnotesize $^*$Flow partition diverges after ep~700: ATT rises from $\sim$15 to
   40--121\,s/veh before partially recovering; late-$\sigma{=}25.3$ indicates
   ongoing oscillation of $\pm$50\,s/veh.}
\end{table}

CoLight$+$METIS ($k{=}3$) is essentially identical to the no-partition baseline
(eval 15.92 vs 16.20, $-1.7\%$): the geometric cut neither helps nor hurts
the global attention mechanism.  CoLight$+$flow, however, suffers \emph{training
instability}: after an initial phase comparable to the other variants
($\sim$15--25\,s/veh through episode~700), ATT grows monotonically to 40--121\,s/veh
by episode~900--1300 before partially recovering, yielding eval ATT of 28.06 (+73\%
vs.\ baseline) and Late-$\sigma{=}25.3$.  This is a divergence failure, not convergence
to a poor optimum.

The mechanism likely involves two compounding factors.
First, with $k{=}3$ on a 7-intersection network, each flow module contains only 2--3
agents: the within-module GAT has insufficient neighbour diversity to aggregate
meaningful representations, and cross-module edges are blocked.
Second, InfoMap partitions are rarely balanced in flow-dominated networks---one module
may attract 5 agents and another only 1---creating unequal attention neighbourhoods
whose gradient contributions are asymmetric and may interfere destructively.
We note that CoLight's global attention may already \emph{discover} flow structure
endogenously; restricting it to pre-specified flow boundaries removes the degrees
of freedom needed to handle cross-module spill-back.

\textbf{Revised belief.}
The partition's benefit appears to be \emph{level-dependent}: flow boundaries reduce
credit-assignment noise at Levels~1--2 (reward and critic), where OD-coherent modules
produce correlated state trajectories.  At Level~3 (attention), pre-specified flow
boundaries restrict the GAT to communication graphs that may be too small or too
unbalanced for effective aggregation on Cologne8.  Whether flow partition is
beneficial at Level~3 on larger networks (Ingolstadt21, where modules would contain
$\approx$4--5 agents) remains an open question for the full study.

% ---- Ingolstadt21 ----

\subsection{Ingolstadt21 Results (1500 Episodes)}
\label{sec:i21prelim}

Eight runs ($2\ \text{algos} \times 2\ \text{partitions} \times 2\ \text{k-values}$,
seed~0) compare MAPPO (Level~2) and IPPO (Level~1) with flow vs.\ METIS at $k{=}4$
and $k{=}5$.  No no-partition baseline has been run yet on this network; results
are therefore \emph{relative to each other}, not to an unpartitioned baseline.
Table~\ref{tab:i21} presents the results ranked by final eval ATT.

\begin{table}[htb]
  \centering
  \caption{Ingolstadt21 results (seed=0, 1500~episodes), eval ATT in s/veh.
           METIS/Flow ratio $=$ METIS$\div$flow ATT at matched (algo, $k$);
           $>1$ means flow partition is better.}
  \label{tab:i21}
  \small
  \begin{tabularx}{\linewidth}{l rr rr rr}
    \toprule
    & \multicolumn{2}{c}{\textbf{Flow}} &
      \multicolumn{2}{c}{\textbf{METIS}} &
      \multicolumn{2}{c}{\textbf{METIS/Flow}} \\
    \cmidrule(lr){2-3}\cmidrule(lr){4-5}\cmidrule(lr){6-7}
    \textbf{Algo} & $k{=}4$ & $k{=}5$ & $k{=}4$ & $k{=}5$ & $k{=}4$ & $k{=}5$ \\
    \midrule
    MAPPO & \textbf{140.96} & 160.51 & 401.17 & 640.82 & 2.85$\times$ & 3.99$\times$ \\
    IPPO  & 564.96 & 202.71 & 846.14 & 526.10 & 1.50$\times$ & 2.60$\times$ \\
    \bottomrule
  \end{tabularx}
\end{table}

\medskip\noindent\textbf{Flow dominates METIS at all conditions.}
At every matched (algo, $k$) pair, the flow partition outperforms METIS:
MAPPO$+$flow $k{=}4$ vs.\ MAPPO$+$METIS $k{=}4$: 141 vs.\ 401\,s/veh
($\mathbf{2.85\times}$ lower ATT); MAPPO$+$flow $k{=}5$ vs.\ METIS $k{=}5$:
161 vs.\ 641\,s/veh ($\mathbf{3.99\times}$); IPPO$+$flow $k{=}5$ vs.\ METIS $k{=}5$:
203 vs.\ 526 ($\mathbf{2.60\times}$); IPPO $k{=}4$: 565 vs.\ 846 ($\mathbf{1.50\times}$).
The advantage is largest for MAPPO, consistent with H1: flow-coherent modules provide
a lower-variance basis for the centralised critic, as within-module agents share
correlated future state trajectories.

\medskip\noindent\textbf{MAPPO $\gg$ IPPO.}
MAPPO$+$flow $k{=}4$ (141\,s/veh) is 4.01$\times$ better than IPPO$+$flow $k{=}4$
(565\,s/veh), and 1.26$\times$ better than IPPO$+$flow $k{=}5$ (203\,s/veh).
The centralized critic (Level~2) amplifies the partition benefit substantially: without
it (IPPO), the flow partition still helps but is more sensitive to $k$ and exhibits
higher training variance.

\medskip\noindent\textbf{METIS instability.}
MAPPO$+$METIS runs do not converge cleanly on Ingolstadt21: late-period standard
deviations exceed 40\,s/veh with frequent spikes above 200\,s/veh.
The $k{=}5$ case (641\,s/veh) is even \emph{worse} than IPPO$+$METIS $k{=}5$
(526\,s/veh), suggesting that METIS boundaries at $k{=}5$ interfere actively with
centralised value-function learning.

\medskip\noindent\textbf{$k$-sensitivity.}
For MAPPO$+$flow, $k{=}4$ outperforms $k{=}5$ by 14\% (141 vs.\ 161),
corresponding to $\approx$5 agents per region---a granularity that appears sufficient
for the centralized critic without making modules too large to coordinate.
For METIS, $k{=}4$ is substantially better than $k{=}5$ for MAPPO (2.4$\times$),
indicating that the geometric cut is more sensitive to over-partitioning.

\subsection{Outstanding Issues and Immediate Actions}

Table~\ref{tab:issues} lists the critical issues identified from the pilot study.

\begin{table}[t]
  \centering
  \caption{Critical issues from pilot study and planned mitigations.}
  \label{tab:issues}
  \small
  \begin{tabularx}{\linewidth}{l X X}
    \toprule
    \textbf{Priority} & \textbf{Issue} & \textbf{Action} \\
    \midrule
    P1 & No no-partition MAPPO/IPPO baseline on Ingolstadt21; cannot
        quantify absolute improvement vs.\ unpartitioned policy &
        Launch no-partition MAPPO/IPPO baselines; report relative-to-baseline
        improvement in final study \\
    P1 & CoLight$+$flow diverges on Cologne8 ($k{=}3$, 7~intersections);
        Level~3 benefit of flow partition unconfirmed for larger networks &
        Run CoLight no-partition + flow on Ingolstadt21 ($k{=}4$,
        $\approx$5~agents/module) to test whether small-module effect explains failure \\
    P2 & IPPO$+$METIS on Cologne8 not converged at ep~1500
        (Late-$\sigma{=}68.5$) &
        Extend run; apply annealing schedule $\alpha(e)$ from
        Eq.~\eqref{eq:alphaanneal} \\
    P2 & MAPPO$+$METIS $k{=}5$ on Ingolstadt21 worse than IPPO$+$METIS $k{=}5$
        (anomalous reversal); cause unclear &
        Investigate whether $k{=}5$ METIS creates a badly connected region
        ($<$3 agents) that destabilises critic training \\
    P3 & Single seed throughout; rliable protocol inapplicable &
        Expand to $\ge$5 seeds for MAPPO$+$flow conditions \\
    \bottomrule
  \end{tabularx}
\end{table}

% ============================================================
\section{Discussion}
\label{sec:discussion}
% ============================================================

\subsection{Anticipated Benefits}

\textbf{Principled coordination graph design.}
The ND-POMDP formulation (Section~\ref{sec:formulation}) establishes that
the optimal coordination graph connects exactly the agent pairs with
non-negligible joint utility.  InfoMap's map-equation optimum
(Eq.~\eqref{eq:mapequation}) provides a principled approximation to this
graph from OD data, without requiring MARL training to discover the structure.
The preliminary results support this: MAPPO with InfoMap partition reaches the
baseline's best ATT level using only 31\% of the training budget, suggesting
that the OD-prior dramatically reduces the number of episodes needed to
discover the effective coordination structure.

\textbf{Who and what are jointly addressed.}
By grounding both the coordination graph (who communicates) and the reward /
value / attention signal (what is communicated) in the same OD flow structure,
the framework provides a \emph{unified} inductive bias---the primary novelty absent
in all prior work.  Preliminary results support this at Levels~1--2: flow partition
reduces credit-assignment noise for MAPPO by up to 4$\times$ relative to geometric
partitioning (METIS) on Ingolstadt21, consistent with H1 (within-module agents share
correlated future state trajectories).  At Level~3 (CoLight attention), the Cologne8
pilot finds that flow boundaries can \emph{hurt} when module size is too small
($k{=}3$, $\approx$2 agents/module); METIS is neutral.  This is an expected
boundary condition rather than a refutation: the map equation has no minimum-module-size
constraint, and with only 7 intersections, $k{=}3$ produces modules too small for
meaningful within-module GAT aggregation.  The full study will test CoLight on
Ingolstadt21 ($k{=}4$, $\approx$5 agents/module), where the hypothesis is that
inductive bias provided by the flow partition will allow the GAT to converge faster by
restricting attention to informationally coherent neighbourhoods rather than the full
graph.  Agents in the same flow basin share future-state MI (H1); their joint reward
is informative (L1); their value function shares structure (L2); and, for sufficiently
large modules, their attention range should concentrate on the coordination-relevant
neighbourhood (L3).  No prior MARL-TSC work addresses all three simultaneously
under a single principled criterion.

\textbf{Spatial dependency is the dominant driver.}
STMARL~\cite{stmarl2020} found that spatial structure dependency boosts TSC
performance more than temporal dependency in most networks.  The InfoMap
partition primarily targets this spatial dimension: it establishes
\emph{who} should coordinate.  The temporal dimension---history windows,
LSTM hidden states---remains handled by the RL policy's own architecture.
This prioritisation is empirically grounded.

\textbf{Scalability.} InfoMap module size grows sub-linearly with network
size: per-agent communication cost is $O(n/k)$ within modules rather than
$O(n)$, and regional critic input dimensionality is $O(n/k)$ rather than
$O(n)$.  The ND-POMDP tractability result~\cite{kok2006ndpomdp} provides
the formal guarantee: a sparser coordination graph with the same coverage
of true dependencies achieves better complexity-quality trade-offs.

\textbf{Interpretability.} Communities correspond to identifiable OD demand
basins that traffic engineers can validate independently of the MARL training
pipeline. The attention range analysis provides an additional mechanistic
visualisation layer for validation.

\textbf{Partition transferability.} The InfoMap partition is recomputable from
any OD flow matrix without MARL training, which is faster than re-running a full
topology learning procedure. The MARL policy still requires retraining on the
new network; partition transferability reduces only the \emph{topology design}
step, not the policy learning step. Zero-shot partition transfer is tested
in Section~\ref{sec:experiment} to bound this claim.

\subsection{Limitations}

Table~\ref{tab:limitations} lists the main threats to validity and their
mitigations.

\begin{table}[t]
  \centering
  \caption{Anticipated limitations and mitigations.}
  \label{tab:limitations}
  \small
  \begin{tabularx}{\linewidth}{l X X}
    \toprule
    \textbf{\#} & \textbf{Limitation} & \textbf{Mitigation} \\
    \midrule
    L1 & Temporal non-stationarity: time-aggregated OD misses peak-hour
        dynamics &
        Time-of-day partitioning ablation (AM/PM/off-peak) \\[4pt]
    L2 & Simulation-reality gap: simplified flow dynamics &
        Multiple network types; real-world topology experiments \\[4pt]
    L3 & OD matrix availability: may not exist in all deployments &
        OD estimation from loop-detector counts is feasible \\[4pt]
    L4 & Module boundary intersections: ambiguous assignment &
        Overlapping partitions (InfoMap soft assignment); explicit
        boundary handling \\[4pt]
    L5 & Strong inter-region flows require cross-module coordination &
        Hierarchical design: intra-module detailed + inter-module summary \\[4pt]
    L6 & Cold-start: warm-up OD $\neq$ converged-policy OD &
        Historical OD preferred; iterative re-partitioning as fallback;
        ablation comparing both strategies \\[4pt]
    L7 & H1/H2 conjecture: flow coherence $\not\Rightarrow$ MI advantage
    or spill-back containment in all scenarios &
    Mandatory pre-training H1+H2 verification; experiment halts if
    both fail \\[4pt]
    L8 & Temporal OD variability: 1-hour static partition may be
    suboptimal for multi-regime episodes &
    Flow stability supports regime-level stability~\cite{flowstability2022};
    temporal ablation (Section~\ref{sec:temporal}) quantifies the gap \\[4pt]
    L9 & ND-POMDP simplification: factored reward assumes separability
    outside $E_C$; real spill-back may create weak cross-module
    coupling &
    Spill-back events measured empirically (H2); cross-module edge
    added when spill-back rate exceeds threshold \\
    \bottomrule
  \end{tabularx}
\end{table}

\subsection{Relationship to Prior Work}

\textbf{Ma \& Wu}~\cite{ma2022feudal} use reward-driven partitioning in a
feudal policy hierarchy. Three distinctions: (i)~their partition criterion
is reward signal (learned from task), ours is OD flow (domain knowledge,
independent of training); (ii)~their partition defines a \emph{policy
  hierarchy}, ours defines a \emph{coordination graph}; (iii)~they provide
no information-theoretic analysis of partition quality.

\textbf{GPLight}~\cite{liu2023gplight} groups agents by current-state
observation MI. Our approach uses future-state propagation (OD flow
dynamics), which is more directly predictive of coordination needs:
the relevant MI for communication is $I(X_i(t); X_j(t+\tau))$, the
future-state signal, not $I(X_i(t); X_j(t))$, the current-state signal.
Furthermore, GPLight's partition is task-dependent and non-transferable;
ours is pre-computed from infrastructure data.

\textbf{STMARL}~\cite{stmarl2020} demonstrates that spatial coordination
dependency dominates temporal dependency in MARL-TSC, motivating the
focus on spatial partition design (InfoMap) rather than temporal context
aggregation. Our three-level protocol isolates the spatial coordination
mechanism across reward, critic, and attention---an experiment STMARL
does not conduct.

\textbf{Dependency Dynamics}~\cite{dependencydyn2025} establishes that
inter-agent coordination is necessary iff spill-back congestion connects
the agents' queues.  Our H2 formalises this as a testable prediction about
InfoMap module boundaries, grounding the partition criterion in traffic
theory rather than only in information theory.  The two works are
complementary: if H2 holds, it provides a mechanistic \emph{causal}
explanation for H1 (why within-module MI is higher: because spill-back
propagates within modules, creating real causal coupling).

\textbf{ExpoComm}~\cite{expocomm2025} demonstrates that pre-designed
domain-grounded topologies can match learned topologies at scale---the
same argument motivates the InfoMap prior.  ExpoComm uses graph-diameter
minimisation; our approach uses flow-coherence maximisation, which is
domain-richer.

The rliable evaluation protocol~\cite{agarwal2021rliable} addresses the
reproducibility concern raised by Henderson et al.~\cite{henderson2018matters}
and is, to our knowledge, not yet standard practice in MARL-TSC evaluation.
The pilot study (Section~\ref{sec:prelim}) illustrates the risk: with $n=1$
seed, IPPO+METIS appears to reach a final ATT of 33.8\,s/veh, but its
Late-$\sigma = 68.5$ reveals the run has not converged---a fact invisible
in any single-seed final-ATT comparison.

% ============================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================

We have presented a research proposal to design MARL communication topologies
for traffic signal control using the map equation (InfoMap).
The proposal makes four interlocking contributions.
First, it reframes the communication design problem using the ND-POMDP
framework: the \emph{coordination graph}---not the road network topology---is
the design variable, and OD flow dynamics provide the principled criterion
for setting its edges.  The map equation's MDL objective formally aligns with
the coordination description length minimisation goal of ND-POMDP, providing
the first theoretical connection between InfoMap and tractable MARL planning.
Second, it operationalises the InfoMap partition at three orthogonal
levels---reward shaping, regional critics, and regional GATs---with two
testable hypotheses (H1: within-module MI advantage; H2: spill-back
containment within modules) and a temporal sensitivity ablation addressing
OD demand stationarity.
Third, it synthesises four literature bodies to identify six research gaps,
the most critical of which---the conflation of road network topology with
coordination graph in all existing MARL-TSC work---motivates the core
algorithmic contribution.
Fourth, it adopts the rliable evaluation framework to ensure statistical
defensibility of sample-efficiency claims.

Preliminary pilot experiments on Cologne8 provide an encouraging directional
signal: MAPPO with InfoMap partition achieves a 10\% reduction in final ATT
and a 43\% reduction in AUC-LC relative to the MAPPO baseline, reaching the
baseline's best performance in 31\% of the training budget, and InfoMap
consistently outperforms METIS in both final performance and training
stability, consistent with H1.  Critical immediate steps are: verifying H1
and H2 on warm-up simulation data, fixing the CoLight implementation (Level~3
validation), launching multi-seed runs with five seeds, and completing the
Ingolstadt21 baseline to enable the rliable statistical protocol.

\medskip\noindent\textbf{AI Disclosure.} This manuscript was developed with
the assistance of AI research and writing tools (Claude, Anthropic) operating
within the ARS deep-research pipeline. All cited sources were retrieved via
web search and verified for existence. The information-theoretic argument in
Sections~\ref{sec:theory} and~\ref{sec:framework} represents a synthesis that
constitutes a testable hypothesis, not an established result.

% ============================================================
\bibliographystyle{refs-style}
\bibliography{refs}

\end{document}
